---
title: "AQI Prediction - Elements of AI Capstone Project"
author: "Filip Škrabák & Patrik Prizbul"
date: "2026-03-15"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(arrow)
library(lubridate)
library(tsibble)
library(here) # for handling paths correctly
library(corrplot)
library(zoo) # for na.approx
library(forecast)
library(generics)
library(fable)
library(feasts)
library(ggtime) ## gg_season
library(tseries) # for adf.test
here::i_am("project.Rmd")
```

# Introduction

The main objective of this project is to predict the Air Quality Index (AQI) for Bratislava, Slovakia, using historical air quality and weather data.
This requires us to merge two datasets - the pollutants dataset and the weather dataset.

## Project Hypothesis: (concept)

Recent pollutant concentrations and weather conditions in Bratislava are jointly sufficient to predict the AQI 24 hours ahead.

# Pollutants Dataset

-   This dataset comes from [European Enviroment Agency (EEA) Air Quality Download Service](https://eeadmz1-downloads-webapp.azurewebsites.net/)
-   It contains pollutants from year 2017 until the end of 2024
-   Dataset type is Primary Validated Data (E1a), which is verified
-   Timezone is stated as UTC+1

First, we load the dataset and take a look at its structure.

```{r}
pollutants_data <- open_dataset(here("pollutants"))
glimpse(pollutants_data)
```

There are 12 columns and 902 260 rows.
First, lets keep only the concerned station (SK0001A), the years of interest (2017-2024) and relevant columns.

```{r}
filtered_air_data <- pollutants_data %>%
  select(Start, Pollutant, Value, Samplingpoint, Validity) %>% # lazy loading
  filter(grepl("SK0001A", Samplingpoint)) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024) %>%
  collect() 
```

Lets look at numeric ranges and unique values of the columns.

```{r}
summary(filtered_air_data)
```

The summary suggests that: - The `Value` column contains some negative values, which are probably sensor errors.
- The `Validity` is either 1 or 2.
According to data source, both are valid, with 2 being below detection limit.
We are keeping both.

Next, we should translate the `Pollutant` codes into human readable names and reshape the data to have columns for each pollutant.
Finally, we make sure the data is sorted by measurement time and we convert it into a tsibble for time series analysis.

```{r}
clean_air_data <- filtered_air_data %>%
  # Translate EU codes into factors
  mutate(Pollutant_Name = case_when(
    Pollutant == 1 ~ "SO2",
    Pollutant == 5 ~ "PM10",
    Pollutant == 7 ~ "O3",
    Pollutant == 8 ~ "NO2",
    #Pollutant == 10 ~ "CO",
    Pollutant == 6001 ~ "PM2.5"
  )) %>%
  select(Start, Pollutant_Name, Value) %>%
  pivot_wider(
    names_from = Pollutant_Name, 
    values_from = Value
  ) %>%
  arrange(Start) %>%
  as_tsibble(index = Start)

glimpse(clean_air_data)
#View(clean_air_data)
```

Now we have a cleaned dataset with 5 pollutants.
Lets check whether our time series has any gaps.
Our data is hourly.

Lets check gaps in our timeseries whether there are any missing hours.
For now, we will

```{r}
gaps <- clean_air_data %>%
  scan_gaps()

nrow(gaps)
gaps
```

There are 410 gaps.
For now, lets fill them with NA values.
Also check the tsibble gaps again to make sure they are filled.

```{r}
clean_air_data <- clean_air_data %>%
  fill_gaps()

has_gaps(clean_air_data)
interval(clean_air_data)
duplicates(clean_air_data)
```

The output confirms there are no more gaps, the interval is 1 hour and there are no duplicates.

```{r}
complete_rows <- sum(complete.cases(clean_air_data)) # 60845
rows <- nrow(clean_air_data) # 70128

missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage

# longest gap
clean_air_data %>%
  as_tibble() %>%
  pivot_longer(-Start, names_to = "pollutant", values_to = "value") %>%
  arrange(pollutant, Start) %>%
  group_by(pollutant) %>%
  mutate(is_na = is.na(value),
         group = cumsum(!is_na & !lag(!is_na, default = TRUE))) %>%
  filter(is_na) %>%
  group_by(pollutant, group) %>%
  summarise(gap_start = min(Start), gap_end = max(Start), hours = n(), .groups = "drop") %>%
  arrange(desc(hours)) %>%
  select(-group) %>%
  mutate(days = hours / 24)

```

13.24% of our data is missing.
The longest gap (sequence of NA values) for any pollutant is 10.2 days for NO2.

Next step is to impute the missing values.
We use a 2 step approach.
- Linear interpolation for small gaps (smaller than 8 hours) - Mean for same hour of day - for larger gaps, this respects the cycles in the data

```{r}
# TODO 
impute_missing_values <- function(x, timestamps, max_linear = 8, window_days = 3) {
  x_out <- zoo::na.approx(x, na.rm = FALSE, maxgap = max_linear)

  hrs <- lubridate::hour(timestamps)
  n   <- length(x_out)

  global_hour_mean <- tapply(x, hrs, mean, na.rm = TRUE)

  for (i in seq_len(n)) {
    if (!is.na(x_out[i])) next

    neighbours <- max(1, i - window_days * 24) : min(n, i + window_days * 24)

    same_hour_and_known <- neighbours[
      hrs[neighbours] == hrs[i] & !is.na(x_out[neighbours])
    ]

    x_out[i] <- if (length(same_hour_and_known) > 0) {
      mean(x_out[same_hour_and_known]) 
    } else {
      global_hour_mean[[ as.character(hrs[i]) ]] 
    }
  }
  x_out
}

complete_air_data <- clean_air_data %>%
  mutate(across(c(SO2, PM10, O3, NO2, `PM2.5`), ~ 
    pmax(impute_missing_values(.x, Start, max_linear = 8, window_days = 3), 0)))

complete_rows <- sum(complete.cases(complete_air_data)) # 60845
rows <- nrow(complete_air_data) # 70128
missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage
```

Now, the dataset is complete with no missing values.

# Weather Dataset

-   This dataset comes from [Open Meteo Historical Weather](https://open-meteo.com/en/docs/historical-weather-api)
-   It contains weather data from year 2017 until the end of 2024
-   Timezone is stated as GMT+0

Loading the dataset and merging it with pollutants:

```{r}
# TODO output formatting
weather_file_path <- here("weather", "open-meteo-48.12N17.10E163m.csv")

# TODO: select more variables, reason about which ones
weather_data <- read_csv(weather_file_path, skip = 3)%>%
  rename(
    Start = time,
    Temperature = `temperature_2m (°C)`,
    Rain = `rain (mm)`,
    Wind_Speed = `wind_speed_10m (km/h)`
  ) %>%
  #select(Start, Temperature, Rain, Wind_Speed) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024)

complete_rows_weather <- sum(complete.cases(weather_data))
rows_weather <- nrow(weather_data)
missing_percentage_weather <- (rows_weather - complete_rows_weather) / rows_weather * 100
missing_percentage_weather

complete_data <- complete_air_data %>%
  inner_join(weather_data, by = "Start")

glimpse(complete_data)
summary(complete_data)
has_gaps(complete_data)
```

The open meteo dataset has no missing values.
The merge has not changed the row count, and our time series has no gaps.
Currently, we have 36 columns.

Next step is to add AQI.

```{r}

# AQI levels and colors
aqi_levels <- c("Good", "Fair", "Moderate", "Poor", "Very poor", "Extremely poor")
aqi_colors <- c("Good"           = "#50f0e6",
                "Fair"           = "#50ccaa",
                "Moderate"       = "#f0e641",
                "Poor"           = "#ff5050",
                "Very poor"      = "#960032",
                "Extremely poor" = "#7d2181")

## European AQI
add_eaqi <- function(df) {
  df %>%
    mutate(
      AQI_PM2.5 = case_when(
        `PM2.5` <=   5 ~ 1L, `PM2.5` <=  15 ~ 2L, `PM2.5` <=  50 ~ 3L,
        `PM2.5` <=  90 ~ 4L, `PM2.5` <= 140 ~ 5L, `PM2.5` >  140 ~ 6L
      ),
      AQI_PM10 = case_when(
        PM10 <=  15 ~ 1L, PM10 <=  45 ~ 2L, PM10 <= 120 ~ 3L,
        PM10 <= 195 ~ 4L, PM10 <= 270 ~ 5L, PM10 >  270 ~ 6L
      ),
      AQI_O3 = case_when(
        O3 <=  60 ~ 1L, O3 <= 100 ~ 2L, O3 <= 120 ~ 3L,
        O3 <= 160 ~ 4L, O3 <= 180 ~ 5L, O3 >  180 ~ 6L
      ),
      AQI_NO2 = case_when(
        NO2 <=  10 ~ 1L, NO2 <=  25 ~ 2L, NO2 <=  60 ~ 3L,
        NO2 <= 100 ~ 4L, NO2 <= 150 ~ 5L, NO2 >  150 ~ 6L
      ),
      AQI_SO2 = case_when(
        SO2 <=  20 ~ 1L, SO2 <=  40 ~ 2L, SO2 <= 125 ~ 3L,
        SO2 <= 190 ~ 4L, SO2 <= 275 ~ 5L, SO2 >  275 ~ 6L
      ),
      AQI = pmax(AQI_PM2.5, AQI_PM10, AQI_O3, AQI_NO2, AQI_SO2, na.rm = TRUE),
      AQI_Label = factor(
        case_when(
          AQI == 1L ~ aqi_levels[1],
          AQI == 2L ~ aqi_levels[2],
          AQI == 3L ~ aqi_levels[3],
          AQI == 4L ~ aqi_levels[4],
          AQI == 5L ~ aqi_levels[5],
          AQI == 6L ~ aqi_levels[6],
        ),
        levels = aqi_levels,
        ordered = TRUE
      )
    )
}

complete_data <- add_eaqi(complete_data)
glimpse(complete_data)
```
As AQI is calculated based on the worst pollutant in the area, we should find out which is the most influencing one (biased by compare order when there are multiple worst pollutants).

```{r}
complete_data %>% 
  as_tibble() %>%
  select(AQI, AQI_SO2, AQI_NO2, AQI_O3, AQI_PM10, AQI_PM2.5) %>%
  mutate(AQI_poll = case_when(
    AQI == AQI_PM2.5 ~ "AQI_PM2.5",
    AQI == AQI_PM10 ~ "AQI_PM10",
    AQI == AQI_O3 ~ "AQI_O3",
    AQI == AQI_NO2 ~ "AQI_NO2",
    AQI == AQI_SO2 ~ "AQI_SO2",
  )) %>%
  group_by(AQI_poll) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

This table shows that the PM2.5 is most commonly the worst pollutant. This is explained by Bratislava's geography, as the city lies in a basin, and it's prone to temperature inversions that trap pollutants. The second worst, NO2 is caused by traffic emissions, and third one, O3 peaks in summer months, as it is formed by photochemical reactions in the presence of sunlight.

This means that predicting AQI is largely equivalent to predicting PM2.5 levels.

# EDA

```{r}
theme_set(theme_minimal() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"))) 
```

## AQI (weekly average)
```{r}
complete_data %>%
  as_tibble() %>%
  mutate(
    Week = floor_date(Start, "week")
  ) %>%
  group_by(Week) %>%
  summarise(AQI = mean(AQI, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(Week, AQI)) +
  geom_line(color = "steelblue", linewidth = 0.4, alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "firebrick", linewidth = 1) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    breaks = 1:6,
    labels = aqi_levels,
    limits = c(1, 6)
  ) +
  labs(
    title = "Weekly average AQI (2017-2024)",
    subtitle = "Red line = LOESS trend",
    x = NULL, y = "AQI level"
  )
```

## AQI Distribution

```{r}
complete_data %>%
  count(AQI_Label) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ggplot(aes(AQI_Label, n, fill = AQI_Label)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", percent)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = aqi_colors) +
  labs(title = "AQI Distribution (2017-2024)", x = "AQI", y = "Count")
```

## Outliers

Boxplots of pollutant concentrations (µg/m³) show which pollutants drive extreme AQI values.

```{r}
complete_data %>%
  as_tibble() %>%
  select(Start, PM2.5, PM10, NO2, O3, SO2) %>%
  pivot_longer(-Start, names_to = "Pollutant", values_to = "Concentration") %>%
  mutate(Pollutant = factor(Pollutant, levels = c("PM2.5", "PM10", "NO2", "O3", "SO2"))) %>%
  ggplot(aes(Pollutant, Concentration, fill = Pollutant)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3, linewidth = 0.4) +
  facet_wrap(~Pollutant, scales = "free_y", nrow = 1) +
  labs(title = "Pollutant distributions with outliers (2017-2024)", x = NULL, y = "Concentration (µg/m³)") +
  theme(legend.position = "none", axis.text.x = element_blank())
```

The plots show that NO2 and PM2.5 have the most frequent outlier hours. SO2, on the other hands, shows the most extreme outliers. As those readings are real world, validated sensor measurements, not errors, they are kept in the dataset.

## Decomposition and seasonality

STL decomposition of AQI:

```{r}
complete_data %>%
  model(STL(AQI ~ season("year") + season("week") + season("day"), robust = TRUE)) %>%
  components() %>%
  autoplot(linewidth = 0.001) +
  labs(title = "STL Decomposition of AQI", x = "Time", y = "AQI") + 
  theme(panel.background = element_rect(),
        panel.grid.major = element_line())

data_season_year <- complete_data %>%
  model(STL(AQI ~ trend() + season("1 year"), robust = TRUE)) %>%
  components()

data_season_year_day <- complete_data %>%
  model(STL(AQI ~ trend() + season("1 year") + season("day"), robust = TRUE)) %>%
  components()

```

```{r}
monthly_aqi <- complete_data %>%
  index_by(Month = yearquarter(Start)) %>%
  summarise(AQI = mean(AQI))

monthly_aqi %>%
  gg_season(AQI, labels = "right") +
  scale_y_continuous(breaks = 1:6,
                     labels = aqi_levels,
                     limits = c(1, 4)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017-2024)")

monthly_aqi %>%
  gg_subseries(AQI) +
  scale_y_continuous(breaks = 1:6,
                     labels = aqi_levels,
                     limits = c(1, 4)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017-2024)")

complete_data %>%
  gg_season(PM2.5, labels = "right") +
  labs(y = "PM2.5", x = element_blank(),
       title = "Yearly PM2.5 (2017-2024)")


```

The plot shows that AQI is worse around winter months (November to March), spring and autumn months have the best air quality, while summer is somewhere in the middle.
This is likely caused by heating in the winter months (by PM2.5).

```{r}
hourly_aqi <- complete_data %>%
  as_tibble() %>%
  mutate(Hour = hour(Start)) %>%
  group_by(Hour) %>%
  summarise(
    Mean  = mean(AQI, na.rm = TRUE),
    Q25   = quantile(AQI, 0.25, na.rm = TRUE),
    Q75   = quantile(AQI, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

hourly_aqi %>%
ggplot(aes(Hour, Mean)) +
  geom_line(color = "steelblue", linewidth = 1) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(
    breaks = 1:6,
    labels = aqi_levels,
    limits = c(1, 6)
  ) +
  labs(
    title = "Hourly AQI pattern (hour of day, 2017-2024)",
    x = element_blank(), y = element_blank()
  )
```

The hourly plot shows that there is a slight peak at around 6am, and an evening peak at around 16pm, or 22pm, likely due to traffic and heating patterns.


## Autocorrelation

### ACF of AQI with all components (2017-2024)

```{r}
complete_data %>%
  ACF(AQI, lag_max = 48 * 7) %>% # up to one week of hourly lags
  autoplot() +
  labs(title = "ACF of AQI (2017-2024)", x = "Lag (hours)", y = "ACF")
```

### ACF of AQI noise - 1 year seasonality removed (STL remainder)

```{r}
data_season_year %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise - 1 year season", x = "Lag (hours)", y = "ACF")
```

### ACF after removing annual + weekly + daily seasonality

```{r}
data_season_year_day %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise (STL: annual + daily removed)", x = "Lag (hours)", y = "ACF")
```

### PACF of AQI

```{r}
complete_data %>%
  PACF(AQI, lag_max = 48) %>%
  autoplot() +
  labs(title = "PACF of AQI (2017-2024)", x = "Lag (hours)", y = "PACF")
```

### Stationarity tests

We run the KPSS and ADF tests to check whether our series is stationary.

```{r warning=FALSE}
tseries::kpss.test(complete_data$AQI)
tseries::kpss.test(noise_season_year$remainder)
tseries::kpss.test(noise_season_year_day$remainder)
```

The KPSS rejects stationarity (H0: stationary), p value \< 0.01

```{r warning=FALSE}
tseries::adf.test(complete_data$AQI)
```

ADF rejects the unit root (H0: series has a unit root), p value \< 0.01

This should indicate seasonal non-stationarity (trend-stationarity).
The mean is not constant, due to winter peaks and summer troughs, but there is no drift in the long run.
Since this time series does not have a unit root, we will not be using differencing as that would likely lead to overdifferencing.

# Random Forest — 24h ahead (minimal prototype)

```{r}
library(ranger)
rf_df <- complete_data %>%
  as_tibble() %>%
  arrange(Start) %>%
  transmute(
    Start,
    AQI_t = AQI,
    AQI_t1 = lag(AQI, 1), # strongest PACF
    AQI_t2 = lag(AQI, 2), # second PACF
    AQI_t24 = lag(AQI, 24), # 24h season
    hour = hour(Start),
    month = month(Start),
    AQI_p24 = lead(AQI, 24) # predicted
  ) %>%
  drop_na() %>%  # remove lead/lag NAs
  mutate(AQI_p24 = factor(AQI_p24, levels = 1:6, ordered = TRUE))


train_rf <- rf_df %>% filter(year(Start) <= 2022) %>% select(-Start)
test_rf  <- rf_df %>% filter(year(Start) >= 2023) %>% select(-Start)

cat("Train rows:", nrow(train_rf), " | Test rows:", nrow(test_rf), "\n")
```

```{r}

# RF (classification, Gini splitting)
set.seed(2026)
rf_model <- ranger(
  AQI_p24 ~ .,
  data = train_rf,
  num.trees = 300,
  importance = "permutation",
  num.threads = parallel::detectCores()
)

rf_model
```

```{r}
# Evaluate on test set
preds_rf <- as.integer(as.character(predict(rf_model, test_rf)$predictions))
actual_rf <- as.integer(as.character(test_rf$AQI_p24))

rmse_rf <- sqrt(mean((actual_rf - preds_rf)^2))
mae_rf <- mean(abs(actual_rf - preds_rf))
acc_rf <- mean(preds_rf == actual_rf)
w1_rf <- mean(abs(preds_rf - actual_rf) <= 1)

cat(sprintf("RMSE: %.4f | MAE: %.4f | Exact: %.1f%% | Within +-1 class: %.1f%%\n", rmse_rf, mae_rf, acc_rf * 100, w1_rf * 100))

tibble(
  Class = 1:6,
  Label = aqi_levels,
  Actual = as.integer(table(factor(actual_rf, levels = 1:6))),
  Predicted = as.integer(table(factor(preds_rf,  levels = 1:6)))
) %>% print()
```

```{r}
# Feature importance
imp <- sort(rf_model$variable.importance,decreasing = TRUE)
barplot(imp, horiz = TRUE, las = 1, cex.names = 0.8,
        main = "RF Permutation Importance (t+24)", xlab = "% IncMSE")
```