---
title: "AQI Prediction - Elements of AI Capstone Project"
author: "Filip Škrabák & Patrik Prizbul"
date: "2026-03-15"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
packages <- c(
  "conflicted", "tidyverse", "arrow", "lubridate", "tsibble",
  "here", # for handling paths correctly
  "corrplot", "zoo", # for na.approx
  "forecast", "generics", "fable",
  "feasts", "ggtime", ## gg_season
  "tseries", # for adf.test
  "MASS", #ordinal regression polr()
  "ranger", "pROC",
  "performance", "brant", "knitr", "scales", "xgboost", "kknn", "vip"
)

missing_packages <- setdiff(packages, rownames(installed.packages()))
if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cran.rstudio.com", dependencies = TRUE)
}

invisible(lapply(packages, library, character.only = TRUE))

conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("interval", "tsibble")
conflict_prefer("vi", "vip")

here::i_am("project.Rmd")
set.seed(2026) # reproducibility
```

# Introduction

The main objective of this project is to predict the Air Quality Index (AQI) for Bratislava, Slovakia, using historical air quality and weather data from Mamateyova street.
This requires us to merge two datasets - the pollutants dataset and the weather dataset. This is a multi-class classification problem, since AQI is categorized into 6 classes, from "Good" to "Extremely poor". Since Bratislava has mostly Fair and Moderate air quality, the dataset is imbalanced, and we have very few examples of Very Poor and Extremely Poor class, which we address by merging AQI into 4 classes (this is explained later).

## Project Hypothesis:

Current and lagged pollutant concentrations, combined with concurrent weather conditions, carry sufficient information to classify the AQI 24 hours ahead in Bratislava with Macro F1 meaningfully above a persistence baseline, with emphasis on correctly identifying the minority Poor class. Weather features provide additional predictive value beyond pollutants alone.

# Pollutants Dataset

-   This dataset comes from [European Enviroment Agency (EEA) Air Quality Download Service](https://eeadmz1-downloads-webapp.azurewebsites.net/)
-   It contains pollutants from year 2017 until the end of 2024
-   Dataset type is Primary Validated Data (E1a), which is verified
-   Timezone is stated as UTC+1

First, we load the dataset and take a look at its structure.

```{r}
pollutants_data <- open_dataset(here("pollutants"))
glimpse(pollutants_data)
```

There are 12 columns and 902 260 rows.
First, lets keep only the concerned station (SK0001A), the years of interest (2017-2024) and relevant columns.

```{r}
filtered_air_data <- pollutants_data %>%
  select(Start, Pollutant, Value, Samplingpoint, Validity) %>% # lazy loading
  filter(grepl("SK0001A", Samplingpoint)) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024) %>%
  collect() 
```

Lets look at numeric ranges and unique values of the columns.

```{r}
summary(filtered_air_data)
```

The summary suggests that: 

- The `Value` column contains some negative values, which are probably sensor errors.
- The `Validity` is either 1 or 2. According to data source, both are valid, with 2 being below detection limit. We are keeping both.

Next, we should translate the `Pollutant` codes into human readable names and reshape the data to have columns for each pollutant.
Finally, we make sure the data is sorted by measurement time and we convert it into a tsibble (time series variant of tibble).

```{r}
clean_air_data <- filtered_air_data %>%
  # Translate EU codes into factors
  mutate(Pollutant_Name = case_when(
    Pollutant == 1 ~ "SO2",
    Pollutant == 5 ~ "PM10",
    Pollutant == 7 ~ "O3",
    Pollutant == 8 ~ "NO2",
    #Pollutant == 10 ~ "CO",
    Pollutant == 6001 ~ "PM2.5"
  )) %>%
  select(Start, Pollutant_Name, Value) %>%
  pivot_wider(
    names_from = Pollutant_Name, 
    values_from = Value
  ) %>%
  arrange(Start) %>%
  as_tsibble(index = Start)

glimpse(clean_air_data)
#View(clean_air_data)
```

Now we have a cleaned dataset with 5 pollutants.
Lets check whether our time series has any gaps.
Our data is hourly.

```{r}
gaps <- clean_air_data %>%
  scan_gaps()

nrow(gaps)
gaps
```

There are 410 gaps.
For now, lets fill them with NA values.
Also check the tsibble gaps again to make sure they are filled.

```{r}
clean_air_data <- clean_air_data %>%
  fill_gaps()

has_gaps(clean_air_data)
interval(clean_air_data)
duplicates(clean_air_data)
```

The output confirms there are no more gaps, the interval is 1 hour and there are no duplicates.

```{r}
complete_rows <- sum(complete.cases(clean_air_data)) # 60845
rows <- nrow(clean_air_data) # 70128

missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage

# longest gap
clean_air_data %>%
  as_tibble() %>%
  pivot_longer(-Start, names_to = "pollutant", values_to = "value") %>%
  arrange(pollutant, Start) %>%
  group_by(pollutant) %>%
  mutate(is_na = is.na(value),
         group = cumsum(!is_na & !lag(!is_na, default = TRUE))) %>%
  filter(is_na) %>%
  group_by(pollutant, group) %>%
  summarise(gap_start = min(Start), gap_end = max(Start), hours = n(), .groups = "drop") %>%
  arrange(desc(hours)) %>%
  select(-group) %>%
  mutate(days = hours / 24)

```

13.24% of our data is missing.
The longest gap (sequence of NA values) for any pollutant is 10.2 days for NO2.

We have dicided that impute the gaps with linear interpolation for small gaps (3 hours). 
Longer gaps will be kept as NA, as replacing them could disrupt the seasonality or trends in data. 
The 3 hour gaps should be short enough to not cause significant bias.

```{r}
complete_air_data <- clean_air_data %>%
  # 0 clamp (pollutant concentrations cannot be negative)
  mutate(across(c(SO2, PM10, O3, NO2, `PM2.5`), ~ pmax(., 0, na.rm = FALSE))) %>%
  mutate(across(c(SO2, PM10, O3, NO2, `PM2.5`), ~ 
    zoo::na.approx(., na.rm = FALSE, maxgap = 3)
    )) 

complete_rows <- sum(complete.cases(complete_air_data)) # 60845
rows <- nrow(complete_air_data) # 70128
missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage
```

Now, after running the linear interpolation for maximum of 3 hour gaps, we reduced the missing percentage to 3.94%, which is acceptable.

# Weather Dataset

-   This dataset comes from [Open Meteo Historical Weather](https://open-meteo.com/en/docs/historical-weather-api)
-   It contains weather data from year 2017 until the end of 2024
-   Timezone is stated as GMT+0

We now load the dataset and rename the columns.

```{r}
weather_file_path <- here("weather", "open-meteo-48.12N17.10E163m.csv")

weather_raw <- read_csv(weather_file_path, skip = 3) %>%
  # We rename the columns manually, as the original names contain spaces and units
  rename(
    Start                 = time,
    Temperature           = `temperature_2m (°C)`,
    Humidity              = `relative_humidity_2m (%)`,
    Dew_Point             = `dew_point_2m (°C)`,
    Apparent_Temp         = `apparent_temperature (°C)`,
    Rain                  = `rain (mm)`,
    Precipitation         = `precipitation (mm)`,
    Snow_Depth            = `snow_depth (m)`,
    Snowfall              = `snowfall (cm)`,
    Weather_Code          = `weather_code (wmo code)`,
    Pressure_MSL          = `pressure_msl (hPa)`,
    Cloud_Cover           = `cloud_cover (%)`,
    Surface_Pressure      = `surface_pressure (hPa)`,
    Cloud_Cover_Low       = `cloud_cover_low (%)`,
    Cloud_Cover_Mid       = `cloud_cover_mid (%)`,
    Cloud_Cover_High      = `cloud_cover_high (%)`,
    ET0                   = `et0_fao_evapotranspiration (mm)`,
    Vapour_Pressure_Def   = `vapour_pressure_deficit (kPa)`,
    Wind_Speed            = `wind_speed_10m (km/h)`,
    Wind_Speed_100m       = `wind_speed_100m (km/h)`,
    Wind_Direction        = `wind_direction_10m (°)`,
    Wind_Gusts            = `wind_gusts_10m (km/h)`,
    Wind_Direction_100m   = `wind_direction_100m (°)`,
    Soil_Temp_0_7         = `soil_temperature_0_to_7cm (°C)`,
    Soil_Temp_7_28        = `soil_temperature_7_to_28cm (°C)`,
    Soil_Temp_28_100      = `soil_temperature_28_to_100cm (°C)`,
    Soil_Temp_100_255     = `soil_temperature_100_to_255cm (°C)`,
    Soil_Moisture_0_7     = `soil_moisture_0_to_7cm (m³/m³)`,
    Soil_Moisture_7_28    = `soil_moisture_7_to_28cm (m³/m³)`,
    Soil_Moisture_28_100  = `soil_moisture_28_to_100cm (m³/m³)`,
    Soil_Moisture_100_255 = `soil_moisture_100_to_255cm (m³/m³)`
  ) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024)

glimpse(weather_raw)
```

## Data Cleaning

```{r}
cat("Rows:", nrow(weather_raw), "| Missing:", sum(!complete.cases(weather_raw)), "\n")
cat("Columns:", ncol(weather_raw), "\n")
```

The weather dataset has no missing values. It has 31 columns. 

We now look at the correlation heatmap to see correlations between weather features.
The heatmap is hierarchically clustered, so the most correlated variables are grouped together, which helps us identify clusters of collinear features.
```{r}
cor_matrix <- weather_raw %>%
  select(-Start, -Weather_Code) %>%
  cor(use = "complete.obs")

cor_matrix %>%
  corrplot(
    method = "color", type = "upper", order = "hclust",
    addCoef.col = "black",
    number.cex = 0.4,
    number.digits = 1,
    tl.cex = 0.55, tl.col = "black",
    col = colorRampPalette(c("steelblue", "white", "firebrick"))(200),
    title = "Weather variable correlations",
    mar = c(0, 0, 1.5, 0)
  )
```


```{r}
cor_matrix %>%
  as_tibble(rownames = "Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation") %>%
  filter(Var1 < Var2) %>%
  arrange(desc(abs(Correlation))) %>%
  head(20)

```

The correlation heatmap confirms that we some variables are extremely highly correlated. Some of there are caused by measuring the same underlying phenomenon, but at different heights (e.g. wind speed at 10m and 100m), or they are calculated from the other columns (e.g. apparent temperature is calculated from temperature, humidity, and wind). Since these features provide us no additional benefit, we drop them.

```{r}
weather_data <- weather_raw %>%
  select(-c(
    Wind_Speed_100m, Wind_Direction_100m, 
    Soil_Temp_7_28, Soil_Temp_28_100, Soil_Temp_100_255,
    Soil_Moisture_7_28, Soil_Moisture_28_100, Soil_Moisture_100_255,
    Surface_Pressure, Apparent_Temp, Snow_Depth, ET0
  ))
```

# Merging the datasets

```{r}
merged_data <- complete_air_data %>%
  inner_join(weather_data, by = "Start")

glimpse(merged_data)
summary(merged_data)
has_gaps(merged_data)
```

After the redundant variables are removed, we retain 24 features in total.
The merge has not changed the row count, and our time series has no gaps.
Both datasets have been successfully merged on the `Start` timestamp, and they have been aligned correctly (no time zone conversion is necessary).

# AQI Calculation

Next step is to add AQI. AQI is calculated based on the worst pollutant in the area, so we calculate AQI level for each pollutant and then take the maximum of those as the overall AQI level. There are 6 AQI levels, from 1 (good) to 6 (extremely poor), based on the concentration thresholds defined by the European Environment Agency (EEA).

```{r}

# AQI levels and colors
aqi_levels <- c("Good", "Fair", "Moderate", "Poor", "Very poor", "Extremely poor")
aqi_colors <- c("Good"           = "#50f0e6",
                "Fair"           = "#50ccaa",
                "Moderate"       = "#f0e641",
                "Poor"           = "#ff5050",
                "Very poor"      = "#960032",
                "Extremely poor" = "#7d2181")

pollutant_colors <- c(
  "PM2.5"       = "#960032",
  "PM10"        = "#ff5050",
  "NO2"         = "#f0e641",
  "O3"          = "#50ccaa",
  "SO2"         = "#50f0e6",
  "Tie/Missing" = "#cccccc"
)

# European AQI calculation
complete_data_six <- merged_data %>%
  mutate(
    AQI_PM2.5 = case_when(
      `PM2.5` <=   5 ~ 1L, `PM2.5` <=  15 ~ 2L, `PM2.5` <=  50 ~ 3L,
      `PM2.5` <=  90 ~ 4L, `PM2.5` <= 140 ~ 5L, `PM2.5` >  140 ~ 6L
    ),
    AQI_PM10 = case_when(
      PM10 <=  15 ~ 1L, PM10 <=  45 ~ 2L, PM10 <= 120 ~ 3L,
      PM10 <= 195 ~ 4L, PM10 <= 270 ~ 5L, PM10 >  270 ~ 6L
    ),
    AQI_O3 = case_when(
      O3 <=  60 ~ 1L, O3 <= 100 ~ 2L, O3 <= 120 ~ 3L,
      O3 <= 160 ~ 4L, O3 <= 180 ~ 5L, O3 >  180 ~ 6L
    ),
    AQI_NO2 = case_when(
      NO2 <=  10 ~ 1L, NO2 <=  25 ~ 2L, NO2 <=  60 ~ 3L,
      NO2 <= 100 ~ 4L, NO2 <= 150 ~ 5L, NO2 >  150 ~ 6L
    ),
    AQI_SO2 = case_when(
      SO2 <=  20 ~ 1L, SO2 <=  40 ~ 2L, SO2 <= 125 ~ 3L,
      SO2 <= 190 ~ 4L, SO2 <= 275 ~ 5L, SO2 >  275 ~ 6L
    ),
    AQI = pmax(AQI_PM2.5, AQI_PM10, AQI_O3, AQI_NO2, AQI_SO2, na.rm = TRUE),
    AQI_Label = factor(
      case_when(
        AQI == 1L ~ aqi_levels[1],
        AQI == 2L ~ aqi_levels[2],
        AQI == 3L ~ aqi_levels[3],
        AQI == 4L ~ aqi_levels[4],
        AQI == 5L ~ aqi_levels[5],
        AQI == 6L ~ aqi_levels[6],
      ),
      levels = aqi_levels,
      ordered = TRUE # AQI is an ordered factor
    )
  )

glimpse(complete_data_six)
```

# Exploratory Data Analysis (EDA)

```{r}
theme_set(theme_minimal() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"))) 
```

## AQI Distribution

```{r}
complete_data_six %>%
  count(AQI_Label) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ggplot(aes(AQI_Label, n, fill = AQI_Label)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", percent)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = aqi_colors) +
  labs(title = "AQI Distribution (2017-2024)", x = "AQI", y = "Count")
```

Air quality in Bratislava is mostly Fair and Moderate (~89%). Good air quality happens in 5.2% of hours. Poor (original EEA classes 4-6, including Very poor and Extremely poor) accounts for 5.7% of hours.

The dataset is imbalanced due to Bratislava having Fair and Moderate air quality most of the time. The Poor class accounts for only 5.7% of the data, yet it is a critical class to be able to predict.
We don't consider this as a problem by itself, as the data distribution reflects the reality of Bratislava's AQI levels, and there are techniques to handle this imbalance.

## Merging Rare AQI Classes

Since Very poor (0.3%) and Extremely poor (0.1%) make a small portion of the dataset, and health warning for general population is issued by Poor level, we merge classes 4 (Poor), 5 (Very poor), and 6 (Extremely poor) into a single "Poor" class. This reduces the problem from 6 to 4 classes: Good, Fair, Moderate, Poor.

```{r}
# Redefine aqi_levels and aqi_colors for the 4 classes
aqi_levels <- c("Good", "Fair", "Moderate", "Poor")
aqi_colors <- c("Good"     = "#50f0e6",
                "Fair"     = "#50ccaa",
                "Moderate" = "#f0e641",
                "Poor"     = "#ff5050")

# Remap AQI
complete_data <- complete_data_six %>%
  mutate(
    AQI = pmin(AQI, 4L),
    AQI_Label = factor(aqi_levels[AQI], levels = aqi_levels, ordered = TRUE)
  )

complete_data %>% 
  as_tibble() %>% 
  count(AQI, AQI_Label) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

The merged "Poor" class accounts for about 5.7% of the data, which is still a minority, but large enough for the models to learn from. We also preserve the health related distinction.
Classes 1 to 3 represent safe conditions, while class 4 signals actionable poor air quality.

## AQI Drivers

As AQI is calculated based on the worst pollutant in the area, we should find out which is the most influencing one. When there are multiple pollutants with the same AQI level, we will label it as "Tie".

```{r}
aqi_drivers <- complete_data %>%
  as_tibble() %>%
  filter(!is.na(AQI)) %>%
  filter(AQI > 1) %>% # "Good" class is the baseline
  mutate(
    n_tied = (AQI_PM2.5 == AQI) + (AQI_PM10 == AQI) +
             (AQI_O3    == AQI) + (AQI_NO2  == AQI) +
             (AQI_SO2   == AQI),
    driver = case_when(
      is.na(n_tied) | n_tied != 1 ~ "Tie/Missing",
      AQI_PM2.5 == AQI ~ "PM2.5",
      AQI_PM10  == AQI ~ "PM10",
      AQI_O3    == AQI ~ "O3",
      AQI_NO2   == AQI ~ "NO2",
      AQI_SO2   == AQI ~ "SO2"
    )
  )

aqi_drivers %>%
  count(driver) %>%
  mutate(Percentage = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n)) %>%
  rename(Pollutant = driver, Hours = n)
```

From the table, we can see that PM2.5 influences the AQI the most, in **20.3%** of cases (caused mainly by heating emissions). The second most influencing pollutant, NO2 (18.2%), is caused by traffic emissions, and the third one, O3 (11.3%), peaks in summer months, as it is formed by photochemical reactions in the presence of sunlight. PM10, which includes dust and coarse particles, is the main driver in 0.4% of hours, and SO2, which is emitted by industries, is the main driver in 0.1% of hours. In 49.7% of hours, there is a tie between multiple pollutants, or the AQI is missing.

```{r}
aqi_drivers %>%
  summarise(n = n(), .by = c(AQI, driver)) %>%
  mutate(pct = 100 * n / sum(n), .by = AQI) %>%
  mutate(
    driver = factor(driver, levels = c("PM2.5", "PM10", "NO2", "O3", "SO2", "Tie/Missing")),
    AQI    = factor(AQI, levels = 1:4, labels = aqi_levels)
  ) %>%
  ggplot(aes(AQI, pct, fill = driver)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = if_else(pct >= 5, sprintf("%.0f%%", pct), "")),
            position = position_stack(vjust = 0.5), size = 3.2, color = "white", fontface = "bold") +
  scale_fill_manual(values = pollutant_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), expand = c(0, 0)) +
  labs(
    title = "Which pollutant drives the AQI? (by class)",
    x = "AQI class", y = "Share of hours (%)", fill = "Pollutant"
  )
```

The Poor class (class 4) is dominated by O3 (37%) and NO2 (31%), with PM2.5 contributing 21%. This is consistent with summer ozone peaks and traffic-related nitrogen dioxide (NO2). 

## Weekly average of AQI across years

```{r}
complete_data %>%
  as_tibble() %>%
  mutate(
    Week = floor_date(Start, "week")
  ) %>%
  group_by(Week) %>%
  summarise(AQI = mean(AQI, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(Week, AQI)) +
  geom_line(color = "steelblue", linewidth = 0.4, alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "firebrick", linewidth = 1) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    breaks = 1:4,
    labels = aqi_levels,
    limits = c(1, 4)
  ) +
  labs(
    title = "Weekly average AQI (2017-2024)",
    subtitle = "Red line = LOESS trend",
    x = NULL, y = "AQI level"
  )
```

The weekly average plot shows a clear repeating seasonal pattern. AQI rises during winter months, and falls during spring and summer. LOESS trend reveals a slight long term improvement in air quality. 

At the beginning of 2017, there is a spike in AQI. This was caused by extremely cold January, which led to increased heating emissions [1].

## Outliers in data

Boxplots of pollutant concentrations (µg/m³) show which pollutants drive extreme AQI values.

```{r}
complete_data %>%
  as_tibble() %>%
  select(Start, PM2.5, PM10, NO2, O3, SO2) %>%
  pivot_longer(-Start, names_to = "Pollutant", values_to = "Concentration") %>%
  mutate(Pollutant = factor(Pollutant, levels = c("PM2.5", "PM10", "NO2", "O3", "SO2"))) %>%
  ggplot(aes(Pollutant, Concentration, fill = Pollutant)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3, linewidth = 0.4) +
  facet_wrap(~Pollutant, scales = "free_y", nrow = 1) +
  labs(title = "Pollutant distributions with outliers (2017-2024)", x = NULL, y = "Concentration (µg/m³)") +
  theme(legend.position = "none", axis.text.x = element_blank())
```

Those readings are real world, validated sensor measurements, not errors. They are kept in the dataset, as they represent the most critical air quality events, which are important to capture.

## Pollutant distributions by AQI class

To understand how individual pollutants relate to AQI, we plot their distributions across the four AQI classes. This should reveal what kind of relationships (linear/nonlinear) we can expect between each pollutant and AQI.

```{r pollutant-by-aqi}
complete_data %>%
  as_tibble() %>%
  select(AQI, PM2.5, NO2, O3, SO2) %>%
  drop_na() %>%
  rename(PM25 = PM2.5) %>%
  mutate(AQI_label = factor(AQI, levels = 1:4, labels = aqi_levels)) %>%
  pivot_longer(c(PM25, NO2, O3, SO2), names_to = "Pollutant", values_to = "Concentration") %>%
  ggplot(aes(AQI_label, Concentration, fill = AQI_label)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, linewidth = 0.4) +
  facet_wrap(~ Pollutant, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = aqi_colors) +
  labs(
    title = "Pollutant concentration by AQI class",
    x = "AQI Class",
    y = "Concentration (µg/m³)",
    fill = "AQI Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 25, hjust = 1),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

PM2.5 and NO2 shows a monotone increase across AQI classes. However, O3's median seems to peak at Fair class, then drops at Moderate and Poor classes. This suggests that there is a non linear relationship between O3 and AQI.

## Decomposition and seasonality

STL decomposition of AQI:

```{r}
complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  model(STL(AQI ~ season("1 year") + season("1 week") + season("1 day"), robust = TRUE)) %>%
  components() %>%
  autoplot(linewidth = 0.001) +
  labs(title = "STL Decomposition of AQI", x = "Time", y = "AQI") + 
  theme(panel.background = element_rect(),
        panel.grid.major = element_line())

data_season_year <- complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  model(STL(AQI ~ trend() + season("1 year"), robust = TRUE)) %>%
  components()

data_season_year_day <- complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  model(STL(AQI ~ trend() + season("1 year") + season("1 day"), robust = TRUE)) %>%
  components()

```
@TODO: opis dekompozicie

```{r}
monthly_aqi <- complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  index_by(Month = yearmonth(Start)) %>%
  summarise(AQI = mean(AQI))
```

```{r}
monthly_aqi %>%
  gg_season(AQI, labels = "right") +
  scale_y_continuous(breaks = 1:4,
                     labels = aqi_levels,
                     limits = c(1, 4)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017-2024)")
```

```{r}
monthly_aqi %>%
  gg_subseries(AQI) +
  scale_y_continuous(breaks = 1:4,
                     labels = aqi_levels,
                     limits = c(1, 4)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017-2024)")
```

The plot shows that AQI is worse around winter months (November to March), spring and autumn months have the best air quality, while summer is somewhere in the middle.
This is likely caused by heating in the winter months (by PM2.5).

```{r}
hourly_aqi <- complete_data %>%
  as_tibble() %>%
  mutate(Hour = hour(Start)) %>%
  group_by(Hour) %>%
  summarise(
    Mean  = mean(AQI, na.rm = TRUE),
    .groups = "drop"
  )

hourly_aqi %>%
ggplot(aes(Hour, Mean)) +
  geom_line(color = "steelblue", linewidth = 1) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(
    breaks = 1:4,
    labels = aqi_levels,
    limits = c(1, 4)
  ) +
  labs(
    title = "Hourly AQI pattern (hour of day, 2017-2024)",
    x = element_blank(), y = element_blank()
  )
```

The hourly plot shows that there is a slight peak at around 6am, and an evening peak at around 16pm, or 22pm, likely due to traffic and heating patterns.


## Autocorrelation

This step should help us understand which lags are the most important for predicting AQI.

### ACF of AQI with all components (2017-2024)

```{r}
complete_data %>%
  ACF(AQI, lag_max = 48 * 7) %>% # up to one week of hourly lags
  autoplot() +
  labs(title = "ACF of AQI (2017-2024)", x = "Lag (hours)", y = "ACF")
```

The plot shows that AQI lags slowly decay. Spikes at 24 hour cycles confirm this daily cycle. There is also a very slight spike at lag 168 (7 days).

### ACF after removing annual + daily seasonality

```{r}
data_season_year_day %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise (STL: annual + daily removed)", x = "Lag (hours)", y = "ACF")
```

Now that both annual and daily components are removed, the lags decay in a much smoother manner. The lag 1 stays high, but the 24 hour spike is gone compared to previous plot. 

### Stationarity tests

We run the KPSS and ADF tests to check whether our series is stationary.

```{r warning=FALSE}
tseries::kpss.test(complete_data$AQI)
tseries::kpss.test(data_season_year_day$remainder)
```

The KPSS rejects stationarity (H0: stationary), p value \< 0.01 for original AQI.

The noise after removing yearly and daily seasonality does not reject null hypothesis.

```{r warning=FALSE}
tseries::adf.test(na.omit(complete_data$AQI))
```

ADF rejects the unit root (H0: series has a unit root), p value \< 0.01

This should indicate seasonal non-stationarity (trend-stationarity).
The mean is not constant, due to winter peaks and summer troughs, but there is no drift in the long run (AQI is mean-reverting). Current values carry genuine information about future values, which justifies using AQI lags as predictive features.

# Feature Engineering

Now, we transform the time series data into a format suitable for machine learning models - we basically make it a supervised learning problem, where the target variable is AQI 24 hours ahead (AQI_p24), and the features are the current and past values of AQI, pollutants, weather variables, and time features (hour, month). This allows us to use many different types of ML models. We decided the lag variables based on autocorrelation analysis.

```{r}
model_features <- complete_data %>%
  as_tibble() %>%
  arrange(Start) %>%
  transmute(
    Start,
    # Target variable
    AQI_p24 = lead(AQI, 24),
    # Autocorrelation lags
    AQI_t   = AQI,
    AQI_t1  = lag(AQI, 1),
    AQI_t2  = lag(AQI, 2),
    AQI_t24 = lag(AQI, 24),
    # Current pollutants
    PM25    = `PM2.5`,
    PM10    = PM10,
    NO2     = NO2,
    O3      = O3,
    SO2     = SO2,
    # Time
    hour    = hour(Start),
    month   = month(Start),
    # Weather
    across(c(Temperature, Humidity, Dew_Point, Rain, Precipitation,
             Snowfall, Weather_Code, Pressure_MSL, Cloud_Cover,
             Cloud_Cover_Low, Cloud_Cover_Mid, Cloud_Cover_High,
             Vapour_Pressure_Def, Wind_Speed, Wind_Direction,
             Wind_Gusts, Soil_Temp_0_7, Soil_Moisture_0_7))
  ) %>%
  drop_na() %>% # remove rows with NA
  # WMO Weather codes are categorical (not ordinal), so we group them into 4 categories based on the weather (we observe 13 codes)
  # This avoids models potentially treating them as ordinal (1-99)
  mutate(Weather_Code = case_when(
    Weather_Code %in% c(0, 1, 2, 3) ~ 1L, # clear / clouds forming
    Weather_Code %in% c(51, 53, 55) ~ 2L, # drizzle
    Weather_Code %in% c(61, 63, 65) ~ 3L, # rain
    Weather_Code %in% c(71, 73, 75) ~ 4L # snow
  ))


# No weather variables (approach 1)
PARTITION_A <- c("AQI_t","AQI_t1","AQI_t2","AQI_t24",
                 "PM25","PM10","NO2","O3","SO2","hour","month")

# All features including weather (approach 2)
PARTITION_B_ALL <- model_features %>%
  select(-Start, -AQI_p24) %>%
  names()

```

## Data Budgeting

```{r}
# Split model_features into train / test by year.
train_data <- model_features %>%
  filter(year(Start) <= 2022) %>%
  select(-Start) %>%
  # AQI_p24 is converted to an ordered factor (1-4) for classification.
  mutate(AQI_p24 = factor(AQI_p24, levels = 1:4, ordered = TRUE))

test_data <- model_features %>%
  filter(year(Start) >= 2023) %>%
  select(-Start) %>%
  mutate(AQI_p24 = factor(AQI_p24, levels = 1:4, ordered = TRUE))

cat("Train rows:", nrow(train_data), " | Test rows:", nrow(test_data), "\n")
```

```{r}
# Class distribution per split (imbalance check)
bind_rows(
  train_data %>% count(AQI_p24) %>% mutate(Split = "Train", pct = round(100 * n / sum(n), 1)),
  test_data  %>% count(AQI_p24) %>% mutate(Split = "Test",  pct = round(100 * n / sum(n), 1))
) %>%
  mutate(Label = aqi_levels[as.integer(AQI_p24)]) %>%
  select(Split, Class = AQI_p24, Label, Count = n, `Pct (%)` = pct) %>%
  arrange(Split, Class)
```

Class 4 (Poor) makes up 6.5% of the training data and 3.3% of the test data.
All four classes are represented, but Poor is a minority class.
We report per-class recall in every evaluation to make this limitation transparent.

## Predictor Correlations

To see possible multicollinearity and relationships between main predictors, we use the pairs plot.
A random sample of 3 000 rows is used (due to performance). We chose a a subset of variables from the dataset, which to our knowledge should be the most important for predicting AQI.

```{r}
# correlation on upper panels
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r     <- cor(x, y, use = "complete.obs")
  r_abs <- abs(r)
  txt   <- format(round(r, digits), nsmall = digits)
  txt   <- paste0(prefix, txt)
  if (missing(cex.cor)) cex.cor <- 0.8 / strwidth(txt)
  col <- if (r >= 0) "firebrick" else "steelblue"
  text(0.5, 0.5, txt, cex = cex.cor * (0.5 + r_abs) / 1.5, col = col)
}

# histogram on the diagonal
panel.hist <- function(x, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5))
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y / max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "steelblue", border = "white")
}

pairs_data <- complete_data %>%
  as_tibble() %>%
  rename(PM25 = `PM2.5`) %>% # for clarity
  select(PM25, PM10, NO2, O3, SO2,
         Temperature, Humidity, Wind_Speed, Pressure_MSL) %>%
  drop_na() %>%
  slice_sample(n = 3000)

pairs(pairs_data,
      lower.panel = panel.smooth,
      upper.panel = panel.cor,
      diag.panel  = panel.hist,
      main = "Pred. relationships and correlations (n = 3 000)")
```

The pairs plot reveals several notable patterns.
PM2.5 and PM10 are strongly correlated (r = 0.85), which indicates that they largely capture the same signal.
NO2 is moderately correlated with PM2.5 and PM10 (r = 0.41), as they all peak during cold, stagnant winter episodes when heating emissions are high and dispersion is low.
O3 is negatively correlated with NO2 (r = -0.57), and does not show a linear relationship with PM2.5/PM10.
Temperature is quite strongly positively correlated with O3 (r = 0.65) as hot, sunny conditions drive ozone formation.
Humidity is strongly negatively correlated with O3 (r = -0.73). Dry and sunny conditions drive high ozone.

## Collinearity Analysis (VIF)

Partition B includes all 18 weather features, on top of pollutants (Partition A). Before any model is fitted, we should check whether we have multicollinearity among the predictors (whether two predictors carry almost identical information). This could make our permutation feature importance in RF unreliable, for example.

We opted to use VIF (Variance Inflation Factor), as it can look at multivariate relationships among features, while correlation only looks at pairwise relationships.

```{r}
library(car)

vif_func <- function(x) {
  y <- seq_len(nrow(x))
  vif(lm(y ~ ., data = x))
}

x_B <- train_data %>% select(all_of(PARTITION_B_ALL))
vif_initial <- vif_func(x_B)

tibble(Feature = names(vif_initial), VIF = round(vif_initial, 1)) %>%
  arrange(desc(VIF)) %>%
  print(n = 30)
```

From the VIF results, we can see that several features have extreme VIF values, three above 1000, and two above 100. Now we take a look at pairwise correlations among the high-VIF features (above 10) to understand this collinearity better.

```{r}
high_vif_features <- c("Rain", "Precipitation", "Snowfall",
                        "Temperature", "Dew_Point", "Soil_Temp_0_7",
                        "Humidity", "Vapour_Pressure_Def",
                        "Wind_Speed", "Wind_Gusts")

cor_hv <- train_data %>%
  select(all_of(high_vif_features)) %>%
  cor(use = "complete.obs")

corrplot(
  cor_hv,
  method = "color", type = "upper", order = "original",
  addCoef.col = "black",
  number.cex = 0.7,
  number.digits = 2,
  tl.cex = 0.75, tl.col = "black",
  col = colorRampPalette(c("steelblue", "white", "firebrick"))(200),
  title = "Correlation for high VIF features",
  mar = c(0, 0, 1.5, 0)
)
```

From the plot, there are three groups of features that are highly correlated with each other.

Precipitation group - Rain and Precipitation are extremely correlated (0.99). Interestingly, snowfall doesn't seem strongly correlated with either of them. This could be because there is not much snowfall in the area.
Temperature group - Temperature, Dew_Point, Soil_Temp_0_7, Humidity, and Vapour_Pressure_Def all measure very similar aspects.
Wind group - Wind_Speed and Wind_Gusts have a strong correlation (r = 0.94).


We also remove some features based on domain knowledge and previous plot or VIF results.

```{r}
# Features removed based on domain knowledge and VIF results
to_remove <- c(
  "Precipitation", # aggregate of Rain + Snowfall
  "Humidity", # relative humidity is temperature-dependent, Dew_Point can capture moisture
  "Vapour_Pressure_Def", # calculated from Temperature + Humidity (redundant)
  "Soil_Temp_0_7", # Dew_Point captures this
  "Temperature", # collinear with Dew_Point (and Dew_Point correlates better with AQI)
  "Wind_Gusts", # captures transient peaks, Wind_Speed captures average 24-hour pollutant dispersion (r = 0.94)
  "PM10"  # collinear with PM2.5 (r = 0.86) and PM2.5 is weighted more heavily (worse health effect)
)
```

Now we run VIF again after removing the 7 features above, to see the max VIF

```{r}
PARTITION_B <- setdiff(PARTITION_B_ALL, to_remove)

x_clean <- train_data %>% select(all_of(PARTITION_B))
vif_final <- vif_func(x_clean)

tibble(
  set     = c("Retained", "Dropped", "Max VIF"),
  value   = c(paste(PARTITION_B, collapse = ", "),
              paste(to_remove, collapse = ", "),
              round(max(vif_final), 2))
) %>% print(n = Inf, width = Inf)
```

We removed 7 features.

The final clean partition has 22 features with max VIF of 4.5, which is acceptable.

# Modelling

All models are trained on 2017-2022 data and evaluated on 2023-2024.
This split ensures no future observations leak into the training data.

## Metrics

We report overall accuracy, +-1 class accuracy, macro-averaged F1 score, and per-class recall. 

We use Macro-Averaged F1 and Poor Class Recall for model evaluation. Macro F1 was chosen because of the class imbalance and the importance of predicting all classes well, not just the majority classes. Model that never predicts Good or Poor is penalised regardless of how well it handles the majority classes. It is important to us that we are able to capture as many as possible of the critical Poor AQI hours, while also maintaining good performance on the other classes.

```{r}
# Macro-averaged F1 (compute F1 per class, then average equally across all 4 classes)
macro_f1 <- function(preds_int, actual_int) {
  f1s <- sapply(1:4, function(k) {
    tp <- sum(preds_int == k & actual_int == k)
    fp <- sum(preds_int == k & actual_int != k)
    fn <- sum(preds_int != k & actual_int == k)
    if (tp == 0) return(0)
    precision <- tp / (tp + fp)
    recall <- tp / (tp + fn)
    2 * precision * recall / (precision + recall)
  })
  round(mean(f1s), 4)
}

# Metrics helper function 
compute_metrics <- function(preds, actual) {
  keep   <- !is.na(actual) # can occur due to lags
  preds  <- preds[keep]
  actual <- actual[keep]
  tibble(
    `Acc (%)` = round(100 * mean(preds == actual), 2),
    `+-1 Class Acc (%)` = round(100 * mean(abs(preds - actual) <= 1), 2),
    `Macro F1` = macro_f1(preds, actual),
    `Poor Recall (%)` = round(100 * mean(preds[actual == 4] == 4), 2),
    `Poor Specificity (%)` = round(100 * mean(preds[actual != 4] != 4), 2)
  )
}

# Recall per class
compute_class_recall <- function(preds, actual) {
  keep   <- !is.na(actual)
  preds  <- preds[keep]
  actual <- actual[keep]
  tibble(Class = 1:4, Label = aqi_levels) %>%
    mutate(
      Actual_n = map_int(Class, \(k) sum(actual == k)),
      Pred_n = map_int(Class, \(k) sum(preds  == k)),
      `Recall (%)` = map_dbl(Class, \(k) {
        n <- sum(actual == k)
        if (n == 0) NA_real_ else round(100 * mean(preds[actual == k] == k), 1)
      })
    )
}

# Confusion matrix heatmap (actual and preds are integer vectors 1-4)
plot_confusion_matrix <- function(actual, preds, title) {
  tibble(Actual = actual, Predicted = preds) %>%
    mutate(
      Actual = factor(Actual, levels = 1:4, labels = aqi_levels),
      Predicted = factor(Predicted, levels = 1:4, labels = aqi_levels)
    ) %>%
    count(Actual, Predicted) %>%
    complete(Actual, Predicted, fill = list(n = 0)) %>%
    mutate(Recall = n / sum(n), .by = Actual) %>%
    ggplot(aes(Predicted, Actual, fill = Recall)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = if_else(n > 0, sprintf("%.0f%%\n(n=%d)", Recall * 100, n), "")),
              size = 2.8, lineheight = 1.2) +
    scale_fill_gradient(low = "white", high = "steelblue",
                        labels = scales::percent, limits = c(0, 1)) +
    scale_x_discrete(position = "top") +
    scale_y_discrete(limits = rev(aqi_levels)) +
    labs(
      title    = title,
      subtitle = "Each cell = % of true class predicted as that column",
      x = "Predicted class", y = "True (Actual) class", fill = "Recall"
    ) +
    theme_minimal()
}
```

## Scenario 1 (Three methods across two feature approaches) + Scenario 2 (3 Parametric and 3 Non-parametric models)

We use the same two feature partitions for all models for a fair comparison of the added value of weather features across different modeling techniques.
Each method is evaluated on two feature sets:

- **Partition A** - AQI lags + current pollutants + time 
- **Partition B** - All features including weather variables

Comparing the two partitions tests whether current weather conditions add predictive value beyond pollutant lags and time features.

We also compare 3 parametric models and 3 non-parametric models (Scenario 2).

### Random Forest

We include the Random Forest (RF) model because it handles non-linear relationships well. During EDA, we observed non-linear relationships between some predictors and AQI (e.g. O3 vs AQI).

Single decision trees are prone to overfitting, but RF builds an ensemble of many trees on random subsets of data and features, which reduces variance and improves generalization.


This model has some limitations, though:
- It cannot extrapolate beyond the range of training data, so it may struggle with extreme values not seen during training.
- It is essentially a "black box", so interpretability is limited compared to simpler models.

We address the class imbalance by using inverse-frequency class weights. During training, misclassifying a minority class (Poor) is penalized more than misclassifying a majority class (Good, Fair, Moderate). This improves recall for Poor AQI hours, which are critical to be able to predict.
Other methods include undersampling/oversampling, but those methods essentially throw away data or create non-existing ones. 

```{r}
library(ranger)

RF_NUM_TREES <- 500 # OOB error stabilises before 500 trees
RF_MAX_DEPTH <- NULL  # NULL = fully grown trees; limit showed no val F1 gain
RF_N_THREADS <- max(1, parallel::detectCores() - 1)
RF_SEED <- 2026
RF_WEIGHT_POWER <- 1.2 # exponent on inverse-frequency weights, tried {0.5, 0.6, ..., 3.0}
# 1.2: best Poor recall / accuracy tradeoff

# other hyperparameters are tuned with grid search (mtry, min_node_size)
```

```{r}
# Computes inverse-frequency case weights for a training set (equivalent to sklearn class_weight="balanced").
# Each sample is weighted by (1 / count_of_its_class)^power.
# power>1 widens the gap between rare and common classes
compute_class_weights <- function(train_data, power = RF_WEIGHT_POWER) {
  class_counts <- table(train_data$AQI_p24)
  weights <- (1 / as.numeric(class_counts))^power
  weights[as.integer(train_data$AQI_p24)]
}

# Fit RF
fit_rf <- function(train_data, partition,
                   num_trees = RF_NUM_TREES,
                   mtry = NULL, # default (sqrt(p)), override with grid value
                   min_node_size = 1, # override with grid value
                   max_depth = RF_MAX_DEPTH,
                   n_threads = RF_N_THREADS,
                   seed = RF_SEED) {

  train_subset <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  weights <- compute_class_weights(train_subset)
  ranger(
    AQI_p24 ~ .,
    data = train_subset,
    num.trees = num_trees,
    mtry = mtry,
    importance = "permutation",
    min.node.size = min_node_size,
    max.depth = max_depth,
    num.threads = n_threads,
    seed = seed,
    case.weights = weights
  )
}

# Returns integer vectors of actual and predicted classes for a fitted RF
predict_rf <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(test_data$AQI_p24)
  preds  <- as.integer(predict(model, data = test_features)$predictions)
  list(actual = actual, preds = preds)
}

library(vip)

eval_rf <- function(model, test_data, partition) {
  p <- predict_rf(model, test_data, partition)
  compute_metrics(p$preds, p$actual)
}

class_recall_rf <- function(model, test_data, partition) {
  p <- predict_rf(model, test_data, partition)
  compute_class_recall(p$preds, p$actual)
}

# bar chart of permutation importance scores
plot_importance <- function(model, title = "RF Permutation Importance") {
  vip(model, num_features = 23, geom = "col") +
    labs(title = title) +
    theme_minimal()
}

# Runs grid search over mtry and min_node_size for a partition
# Returns a tibble with train_f1, val_f1, val_acc and overfitting gap
run_rf_grid <- function(partition) {
  grid <- expand.grid(
    mtry = c(2, 3, 5, 7, 10),
    min_node_size = c(1, 3, 5, 10, 20)
  )

  pmap(grid, function(mtry, min_node_size) {
    m <- fit_rf(rf_train_grid, partition,
                   mtry = mtry, min_node_size = min_node_size)
    p_tr <- predict_rf(m, rf_train_grid, partition)
    p_v <- predict_rf(m, rf_val_data,   partition)

    tibble(
      mtry = mtry,
      min_node_size = min_node_size,
      train_f1 = macro_f1(p_tr$preds, p_tr$actual),
      val_f1 = macro_f1(p_v$preds,  p_v$actual),
      val_acc = round(100 * mean(p_v$preds == p_v$actual), 2),
      gap = train_f1 - val_f1
    )
  }) %>% list_rbind()
}

# Best combo - highest val_f1, then smallest overfitting gap, then min_node_size
best_rf_combo <- function(grid_results) {
  grid_results %>%
    arrange(desc(val_f1), gap, min_node_size) %>%
    slice(1)
}
```

#### Grid Search

`mtry` - number of features randomly considered at each split. Lower values increase tree diversity. Default is square root of number of features.
`min.node.size` - minimum number of samples required to form a leaf. Larger values produce shallower trees, reducing variance and improving minority-class generalization.
`num.trees` - fixed at 500, as adding more trees in RF never causes overfitting (only reduces variance), so it does not need to be grid-searched. OOB error stabilises well before 500.

We sweep `mtry` for {2, 3, 5, 7, 10} and `min.node.size` for {1, 3, 5, 10, 20} separately for each partition, using a temporal split (train <= 2021, validate 2022). Both partitions use inverse-frequency class weights.

```{r rf-grid, cache=TRUE}
# temporal split
rf_train_grid <- model_features %>% filter(year(Start) <= 2021)
rf_val_data   <- model_features %>% filter(year(Start) == 2022)

#grid_A <- run_rf_grid(PARTITION_A)
#grid_B <- run_rf_grid(PARTITION_B)

#best_A <- best_rf_combo(grid_A)
#best_B <- best_rf_combo(grid_B)

#cat("=== Best Partition A ===\n"); print(best_A)
#cat("\n=== Best Partition B ===\n"); print(best_B)
```

From the grid search results, we can see the results for each partition - which hyperparameters achieved the best validation F1, and what was the overfitting gap (train F1 - val F1).

For Partition B, `mtry = 3` and `min_node_size = 20` achieves the best validation macro F1 (0.322) with the lowest overfitting gap (0.070). Higher `mtry` values inflate training F1 without improving validation performance. For Partition A (pollutants only), the optimal is `mtry = 5` and `min_node_size = 20`.

```{r}
# Set per-partition hyperparameters from grid results
RF_MTRY_A <- 5 # best_A$mtry
RF_MIN_NODE_SIZE_A <- 20 # best_A$min_node_size
RF_MTRY_B <- 3 # best_B$mtry
RF_MIN_NODE_SIZE_B <- 20 # best_B$min_node_size
```

```{r fit-rf, cache=TRUE}
rf_A <- fit_rf(train_data, PARTITION_A, mtry = RF_MTRY_A, min_node_size = RF_MIN_NODE_SIZE_A)
rf_B <- fit_rf(train_data, PARTITION_B, mtry = RF_MTRY_B, min_node_size = RF_MIN_NODE_SIZE_B)
```

Overall performance on the 2023-2024 test set:

```{r}
evals_RF <- list(
  "A - pollutants only" = eval_rf(rf_A, test_data, PARTITION_A),
  "B - all features" = eval_rf(rf_B, test_data, PARTITION_B)
)

evals_RF %>%
  list_rbind(names_to = "Partition")
```

OOB F1 (each row predicted only by trees that never saw it) matches test F1 (no overfitting)

```{r}
oob_preds <- rf_B$predictions
valid_oob  <- !is.na(oob_preds)

tibble(
  set = c("Train in-sample (biased)", "OOB (unbiased)", "Test 2023-24"),
  macro_f1 = c(
    macro_f1(as.integer(predict(rf_B, data = train_data %>% select(all_of(c(PARTITION_B, "AQI_p24"))))$predictions), as.integer(train_data$AQI_p24)),
    macro_f1(as.integer(oob_preds[valid_oob]), as.integer(train_data$AQI_p24[valid_oob])),
    eval_rf(rf_B, test_data, PARTITION_B)$`Macro F1`
  )
)
```

Per-class recall:
```{r}
cat("--- Partition A ---\n")
class_recall_rf(rf_A, test_data, PARTITION_A) %>% print()
cat("\n--- Partition B ---\n")
class_recall_rf(rf_B, test_data, PARTITION_B) %>% print()
```

```{r}
plot_importance(rf_A, "RF Permutation Importance (Partition A)")
```

```{r}
plot_importance(rf_B, "RF Permutation Importance (Partition B)")
```

#### Confusion matrix

```{r}
preds_B <- predict_rf(rf_B, test_data, PARTITION_B)
actual_cls <- preds_B$actual
pred_cls <- preds_B$preds
act_pred_tb <- tibble(Actual = actual_cls, Predicted = pred_cls)

plot_confusion_matrix(actual_cls, pred_cls, "Confusion Matrix: RF Partition B")
```

The confusion matrix shows recall per class on the diagonal.
The model is strong for Fair (71.8%) and Moderate (49.1%). With class weighting (power 1.2), Poor recall rises to **53.3%**.
Most misclassifications are adjacent-class (±1) errors (for example, Poor predicted as Moderate, or Good predicted as Fair).

#### Predicted vs. actual class distribution

```{r}
act_pred_tb %>%
  pivot_longer(c(Actual, Predicted), names_to = "Type", values_to = "Class") %>%
  count(Type, Class) %>%
  mutate(Label = factor(aqi_levels[Class], levels = aqi_levels)) %>%
  complete(Type, Label, fill = list(n = 0)) %>%
  ggplot(aes(Label, n, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Actual" = "steelblue", "Predicted" = "tomato")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Predicted vs. actual class distribution (test set 2023-2024)",
    x = NULL, y = "Count", fill = NULL
  ) +
  theme_minimal()
```

The model over-predicts Fair and under-predicts Good and Moderate.

#### One vs Rest ROC and AUC

One VS Rest ROC (OvR) allows us to see how tell the model separates each class from the others.
The `ranger` package doesn't provide probability estimates without `probability = TRUE`, so we opted to train a separate probability forest for this evaluation. All parameters and hyperparameters are kept the same as the original RF to ensure a fair comparison.

```{r}
# Probability forest
rf_B_prob <- ranger(
  AQI_p24 ~ .,
  data = train_data %>% select(all_of(c(PARTITION_B, "AQI_p24"))),
  num.trees = RF_NUM_TREES,
  mtry = RF_MTRY_B,
  min.node.size = RF_MIN_NODE_SIZE_B,
  num.threads = RF_N_THREADS,
  seed = RF_SEED,
  case.weights = compute_class_weights(train_data),
  probability = TRUE # outputs probabilities instead of class labels
)
```

```{r}
library(pROC)

prob_matrix <- predict(rf_B_prob, data = test_data)$predictions

# OVR ROC
roc_list <- map(1:4, function(k) { # loop over aqi
  binary <- as.integer(actual_cls == k) # 1 for class k, 0 for others
  if (length(unique(binary)) < 2) return(NULL)   # skip if no positives in test
  roc(binary, prob_matrix[, k], quiet = TRUE)
})

# tidy data frame of all ROC curves for plotting (for ggplot)
# bit more complex, but allows us to use more pretty ggplot with ordering
roc_df <- map(1:4, function(k) {
  if (is.null(roc_list[[k]])) return(NULL)
  coords_df <- pROC::coords(roc_list[[k]], "all", ret = c("fpr", "tpr")) # get FPR and TPR for all thresholds
  coords_df$class <- aqi_levels[k] # label for the class
  coords_df$auc <- round(pROC::auc(roc_list[[k]]), 3) # AUC for the class for legend
  coords_df
}) %>% list_rbind()

roc_df %>%
  mutate(label = paste0(class, "  (AUC=", auc, ")"), # label for the legend
         label = factor(label, levels = unique(label[order(auc, decreasing = TRUE)]))) %>%
  ggplot(aes(fpr, tpr, color = label)) +
  geom_line(linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_brewer(palette = "Dark2") + # color palette
  labs(
    title = "One VS Rest ROC (RF Partition B weighted)",
    x = "False Positive Rate", y = "True Positive Rate", color = "Class (AUC)"
  ) +
  theme_minimal()
```

The strongest class is **Poor (AUC = 0.889)**, meaning the model reliably separates poor air quality from safe conditions.
Good is also well separated (AUC = 0.776). Moderate and Fair, being the dominant classes, are harder to distinguish from each other: AUC = 0.728 and 0.700, respectively.

### XGBoost

This ensemble method builds trees sequentially, as opposed to RF which builds trees independently in parallel. New trees correct the residual errors of all previous trees combined. This allows XGBoost to learn from its mistakes at each step. The justification for selecting this model is the same as for RF, that there are non linear relationships between predictors and AQI, which linear models cannot capture. In addition, boosting can reduce systematic bias, which is likely present at the Moderate/Poor boundary due to borderline observations that are hard to classify. By iteratively focusing on residual errors, boosting can improve performance on these hard cases, which may improve recall for the critical Poor class.

We include the same inverse-frequency class weights as in RF to address class imbalance. XGBoost thus serves as a strong alternative to RF, allowing direct comparison.

```{r}
library(xgboost)

XGB_BOOSTER <- "gbtree" # for structured data
XGB_OBJECTIVE <- "multi:softmax" # multi-class classification with integer labels

XGB_ETA <- 0.1 # learning rate
XGB_SUBSAMPLE <- 0.8 # 80% of rows per tree
XGB_COLSAMPLE <- 0.8 # 80% of features per tree
XGB_WEIGHT_POWER <- 0.6 # exponent on inverse-frequency weights (same logic as RF)
XGB_SEED <- 2026
XGB_NTHREAD <- max(1, parallel::detectCores() - 1)
```

#### Hyperparameter tuning - Grid search

Key hyperparameters for XGBoost include:

- `nrounds`: number of boosting rounds (trees). More rounds can improve performance but risk overfitting.
- `max_depth`: maximum depth of each tree. Shallower trees = weaker learners, less overfitting.
- `min_child_weight`: minimum number of samples required in a leaf node. Higher values force coarser splits, reducing overfitting on rare classes.
- `eta` (learning rate): step size at each boosting iteration. Smaller eta requires more rounds but reduces overfitting.
- `subsample` and `colsample_bytree`: fractions of rows and features sampled per tree.

We use grid search (train on <= 2021, validate on 2022) to find the best combination of `nrounds`, `max_depth`, and `min_child_weight`.

Grid search sweeps:
- `max_depth` {3, 4, 5, 6}
- `nrounds` {100, 200, 300, 500}
- `min_child_weight` {1, 10}

```{r}
# converts the tibble to a dmatrix
create_dmatrix <- function(data, partition) {
  X <- data %>% 
    select(all_of(partition)) %>% 
    as.matrix()

  # AQI_p24 is ord (1-4), XGB requires 0 based int labels
  y <- as.integer(data$AQI_p24) - 1L

  xgb.DMatrix(X, label = y)
}

dtrain_A <- create_dmatrix(train_data, PARTITION_A)
dtrain_B <- create_dmatrix(train_data, PARTITION_B)
dtest_A <- create_dmatrix(test_data,  PARTITION_A)
dtest_B <- create_dmatrix(test_data,  PARTITION_B)

xgb_params <- list(
  booster = XGB_BOOSTER,
  objective = XGB_OBJECTIVE,
  num_class = 4L,
  eta = XGB_ETA,
  subsample = XGB_SUBSAMPLE,
  colsample_bytree = XGB_COLSAMPLE,
  eval_metric = "mlogloss",
  nthread = XGB_NTHREAD,
  seed = XGB_SEED
)

# inverse-frequency class weights (XGBoost normalized)
compute_class_weights_xgb <- function(data, power) {
  class_counts <- table(data$AQI_p24)
  w_raw <- (1 / as.numeric(class_counts))^power
  w_per_row <- w_raw[as.integer(data$AQI_p24)]
  w_per_row / mean(w_per_row)
}
```

```{r xgboost-grid, cache=TRUE}
train_grid <- model_features %>% filter(year(Start) <= 2021)
val_data <- model_features %>% filter(year(Start) == 2022)

param_grid <- expand.grid(
  max_depth = c(3, 4, 5, 6),
  nrounds = c(100, 200, 300, 500),
  min_child_weight = c(1, 10)
)

run_xgb_grid <- function(partition) {
  dval_gs <- create_dmatrix(val_data, partition)
  dtrain_grid <- create_dmatrix(train_grid, partition)
  w_grid <- compute_class_weights_xgb(train_grid, XGB_WEIGHT_POWER)
  setinfo(dtrain_grid, "weight", w_grid)

  pmap(param_grid, function(max_depth, nrounds, min_child_weight) {
    p <- modifyList(xgb_params, list(max_depth = max_depth, min_child_weight = min_child_weight))
    m <- xgb.train(params = p, data = dtrain_grid, nrounds = nrounds, verbose = 0)

    preds_tr <- as.integer(predict(m, dtrain_grid)) + 1L
    preds_v  <- as.integer(predict(m, dval_gs)) + 1L

    tibble(
      max_depth = max_depth,
      nrounds = nrounds,
      min_child_weight = min_child_weight,
      train_f1 = macro_f1(preds_tr, as.integer(train_grid$AQI_p24)),
      val_f1 = macro_f1(preds_v,  as.integer(val_data$AQI_p24)),
      val_acc = round(100 * mean(preds_v == as.integer(val_data$AQI_p24)), 2),
      gap = train_f1 - val_f1
    )
  }) %>% list_rbind()
}

#grid_A <- run_xgb_grid(PARTITION_A)
#grid_B <- run_xgb_grid(PARTITION_B)

#cat("=== Best Partition A ===\n")
#best_A <- grid_A %>% arrange(desc(val_f1), gap, nrounds, max_depth) %>% slice(1) %>% print()
#cat("\n=== Best Partition B ===\n")
#best_B <- grid_B %>% arrange(desc(val_f1), gap, nrounds, max_depth) %>% slice(1) %>% print()
```

```{r}
# Set per-partition hyperparameters from grid results (mirrors RF structure)
XGB_MAX_DEPTH_A <- 4  # best_A$max_depth
XGB_MIN_CHILD_WEIGHT_A <- 1  # best_A$min_child_weight
XGB_NROUNDS_A <- 200 # best_A$nrounds

XGB_MAX_DEPTH_B <- 6  # best_B$max_depth
XGB_MIN_CHILD_WEIGHT_B <- 1  # best_B$min_child_weight
XGB_NROUNDS_B <- 200 # best_B$nrounds

xgb_params_A <- modifyList(xgb_params, list(max_depth = XGB_MAX_DEPTH_A, min_child_weight = XGB_MIN_CHILD_WEIGHT_A))
xgb_params_B <- modifyList(xgb_params, list(max_depth = XGB_MAX_DEPTH_B, min_child_weight = XGB_MIN_CHILD_WEIGHT_B))
```

We also calculate the overfitting gap (train F1 - val F1) for each combination. The best hyperparameters per partition are printed above and applied to `xgb_params_A` / `xgb_params_B` for the final models.

#### Fitting the model

```{r fit-xgb, cache=TRUE}
fit_xgb <- function(dtrain, nrounds, params) {
  w <- compute_class_weights_xgb(train_data, power = XGB_WEIGHT_POWER)
  setinfo(dtrain, "weight", w)
  xgb.train(params = params, data = dtrain, nrounds = nrounds, verbose = 0)
}

predict_xgb <- function(model, dtest, test_data) {
  preds  <- as.integer(predict(model, dtest)) + 1L  # XGBoost outputs 0-based labels
  actual <- as.integer(test_data$AQI_p24)
  list(actual = actual, preds = preds)
}

eval_xgb <- function(model, dtest, test_data) {
  p <- predict_xgb(model, dtest, test_data)
  compute_metrics(p$preds, p$actual)
}

class_recall_xgb <- function(model, dtest, test_data) {
  p <- predict_xgb(model, dtest, test_data)
  compute_class_recall(p$preds, p$actual)
}

xgb_A <- fit_xgb(dtrain_A, nrounds = XGB_NROUNDS_A, params = xgb_params_A)
xgb_B <- fit_xgb(dtrain_B, nrounds = XGB_NROUNDS_B, params = xgb_params_B)
```

#### Evaluation

```{r}
list(
  "A - pollutants only" = eval_xgb(xgb_A, dtest_A, test_data),
  "B - all features" = eval_xgb(xgb_B, dtest_B, test_data)
) %>%
  list_rbind(names_to = "Partition")

cat("--- Partition A ---\n")
class_recall_xgb(xgb_A, dtest_A, test_data) %>% print()
cat("\n--- Partition B ---\n")
class_recall_xgb(xgb_B, dtest_B, test_data) %>% print()
```

Partition B (all features) achieves higher Macro F1 than Partition A (pollutants only). Class weighting significantly improves recall for the Poor class at the cost of overall accuracy.

#### Confusion matrix

```{r}
p_xgb <- predict_xgb(xgb_B, dtest_B, test_data)
plot_confusion_matrix(p_xgb$actual, p_xgb$preds, "Confusion Matrix: XGBoost Partition B")
```

#### Feature importance

```{r}
xgb_pred_wrapper <- function(object, newdata) {
  dm <- xgb.DMatrix(as.matrix(newdata))
  preds <- as.integer(predict(object, dm)) + 1L
  factor(as.character(preds), levels = c("1", "2", "3", "4"))
}

# XGBoost permutation importance (on test set)
set.seed(XGB_SEED)
xgb_imp_perm_B <- vi(
  object = xgb_B,
  method = "permute",
  train = test_data %>% select(all_of(PARTITION_B)),
  target = test_data$AQI_p24,
  metric = "accuracy",
  pred_wrapper = xgb_pred_wrapper,
  nsim = 5
)

vip(xgb_imp_perm_B, num_features = 30, geom = "col") +
  labs(
    title = "Permutation-based Variable Importance",
    subtitle = "Accuracy metric, 5 permutations"
  ) +
  theme_minimal()

print(xgb_imp_perm_B, n = 30)
```

`Pressure_MSL` is the most dominant feature, nearly 2x the next-ranked feature. High pressure traps pollutants, while low pressure disperses them. `O3` and `PM25` are the next top features.
`Snowfall` has slightly negative importance, meaning shuffling it marginally improves accuracy.

### k-Nearest Neighbours (kNN)

kNN is a distance based method, which uses Euclidean distance to find the k closest training observations to each test observation, then classifies the test observation based on the majority class among those neighbors.
kNN has the assumption that similar points can be found near one another.

This model assumes meaningful distances between feature vectors, so all numeric features are z-score scaled to ensure no single feature dominates the distance calculation.

```{r}
library(kknn) # weighted kNN

KNN_DISTANCE <- 2 # p=2 = Euclidean distance
KNN_TUNE_N <- 5000 # training rows used for LOO-CV tuning
KNN_KMAX <- 100 # upper bound for k search
```

#### Hyperparameter Tuning

We do LOOCV (Leave One Out Cross Validation) for k = 1 to KNN_KMAX to find the optimal k and kernel. 
```{r}
# subsample for tuning
set.seed(2026)
tune_sample_idx <- sample(nrow(train_data), KNN_TUNE_N)

# kernels to try
kernels <- c("rectangular", "triangular", "epanechnikov")

# tuning
tune_knn_cv <- function(partition) {
  s <- train_data[tune_sample_idx, ] %>% select(all_of(c(partition, "AQI_p24")))
  train.kknn(AQI_p24 ~ ., data = s, kmax = KNN_KMAX,
             kernel = kernels,
             distance = KNN_DISTANCE, scale = TRUE)
}

cv_knn_A <- tune_knn_cv(PARTITION_A)
cv_knn_B <- tune_knn_cv(PARTITION_B)

```

```{r}
KNN_K_A      <- cv_knn_A$best.parameters$k
KNN_KERNEL_A <- cv_knn_A$best.parameters$kernel
KNN_K_B      <- cv_knn_B$best.parameters$k
KNN_KERNEL_B <- cv_knn_B$best.parameters$kernel

cat("Best A: k=", KNN_K_A, " kernel=", KNN_KERNEL_A, "\n")
cat("Best B: k=", KNN_K_B, " kernel=", KNN_KERNEL_B, "\n")
```

The triangular kernel wins for both partitions, with k = 30 for Partition A and k = 20 for Partition B.

The triangular kernel assigns higher weights to closer neighbors, and lower weights to farther neighbors.

#### Running KNN and Evaluation

Since KNN is a lazy learner, there is no separate fit (training) step, and it stores no parameters. It requires training set at inference time.

```{r}
fit_knn <- function(train_data, test_data, partition, k, kernel,
                    distance = KNN_DISTANCE) {
  train_sub <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  test_sub  <- test_data  %>% select(all_of(c(partition, "AQI_p24")))
  kknn(AQI_p24 ~ .,
       train = train_sub,
       test  = test_sub,
       k = k,
       distance = distance,
       kernel = kernel,
       scale = TRUE)  # scaling due to kNN sensitivity
}

predict_knn <- function(knn_fit, test_data) {
  preds  <- as.integer(fitted(knn_fit))
  actual <- as.integer(test_data$AQI_p24)
  list(actual = actual, preds = preds)
}

eval_knn <- function(knn_fit, test_data) {
  p <- predict_knn(knn_fit, test_data)
  compute_metrics(p$preds, p$actual)
}

class_recall_knn <- function(knn_fit, test_data) {
  p <- predict_knn(knn_fit, test_data)
  compute_class_recall(p$preds, p$actual)
}
```

```{r}
knn_A <- fit_knn(train_data, test_data, PARTITION_A, k = KNN_K_A, kernel = KNN_KERNEL_A)
knn_B <- fit_knn(train_data, test_data, PARTITION_B, k = KNN_K_B, kernel = KNN_KERNEL_B)

list(
  "A - pollutants only" = eval_knn(knn_A, test_data),
  "B - all features"    = eval_knn(knn_B, test_data)
) %>%
  list_rbind(names_to = "Partition")

cat("--- Partition A ---\n")
class_recall_knn(knn_A, test_data) %>% print()
cat("\n--- Partition B ---\n")
class_recall_knn(knn_B, test_data) %>% print()
```

Partition B overperforms A in Macro F1 slightly (0.334 vs 0.369), which again suggests that weather features add predictive value beyond pollutant lags.
Both partitions achieve >= 97% +-1 class accuracy, meaning that when the model is wrong it is almost always off by just one AQI category.

Compared to RF, kNN does not offer class weight parameter. This means that kNN is affected by class imbalance more severely, as the imbalance directly shapes its predictions (model almost never outputs Good or Poor).

We also tried oversampling the training data, which helped for Poor recall, but overall accuracy dropped significantly (to 47.1%), along with +-1 accuracy. 
We opted not to include that model, as the trade off is worse than the class weighting approach used for RF.

#### Confusion matrix

```{r}
p_knn <- predict_knn(knn_B, test_data)
plot_confusion_matrix(p_knn$actual, p_knn$preds, "Confusion Matrix: kNN Partition B")
```

### Ordinal regression

```{r}
ORDINAL_WEIGHT_POWER <- 0.2 # 0 = no reweighting, 1 = fully balanced

# polr() treats weights as frequency weights (each row counts as w observations)
# Weights must sum to nrow(train_subset) so the effective sample size equals the actual sample size.
compute_class_weights_ordinal <- function(train_data, power = ORDINAL_WEIGHT_POWER) {
  class_counts <- table(train_data$AQI_p24)
  w_raw     <- (1 / as.numeric(class_counts))^power
  w_per_row <- w_raw[as.integer(as.character(train_data$AQI_p24))]
  # Normalize so weights sum to n (preserves effective sample size)
  w_per_row / sum(w_per_row) * nrow(train_data)
}

fit_ordinal <- function(train_data, partition) {
  train_subset_ordinal <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  polr(AQI_p24 ~ ., data = train_subset_ordinal, Hess = TRUE,
       weights = compute_class_weights_ordinal(train_subset_ordinal),
       method = "logistic")
}

ordinal_A <- fit_ordinal(train_data, PARTITION_A)
ordinal_B <- fit_ordinal(train_data, PARTITION_B)
```

#### Coefficient table with p-values

`polr` only outputs t-values; p-values must be derived from the normal approximation (valid for large n).

```{r}
polr_summary_table <- function(model) {
  tbl <- as.data.frame(coef(summary(model)))
  tbl$p_value <- pnorm(abs(tbl[, "t value"]), lower.tail = FALSE) * 2
  round(tbl[, c("Value", "Std. Error", "t value", "p_value")], 4)
}

cat("=== Partition A ===\n"); print(polr_summary_table(ordinal_A))
cat("\n=== Partition B ===\n"); print(polr_summary_table(ordinal_B))
```

#### Performance evaluation

The overall perfomrance and class recall metrics.
```{r}

predict_ordinal <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(as.character(test_data$AQI_p24))
  preds  <- as.integer(as.character(predict(model, newdata = test_features)))
  list(actual = actual, preds = preds)
}

eval_ordinal <- function(model, test_data, partition) {
  p <- predict_ordinal(model, test_data, partition)
  compute_metrics(p$preds, p$actual)
}

class_recall_ordinal <- function(model, test_data, partition) {
  p <- predict_ordinal(model, test_data, partition)
  compute_class_recall(p$preds, p$actual)
}

list(
  "A - pollutants only" = eval_ordinal(ordinal_A, test_data, PARTITION_A),
  "B - all features"    = eval_ordinal(ordinal_B, test_data, PARTITION_B)
) %>%
  list_rbind(names_to = "Partition")

cat("--- Partition A ---\n")
class_recall_ordinal(ordinal_A, test_data, PARTITION_A) %>% print()
cat("\n--- Partition B ---\n")
class_recall_ordinal(ordinal_B, test_data, PARTITION_B) %>% print()
```

#### Confusion matrix

```{r}
p_ord <- predict_ordinal(ordinal_B, test_data, PARTITION_B)
plot_confusion_matrix(p_ord$actual, p_ord$preds, "Confusion Matrix: Ordinal Regression Partition B")
```

#### Assumption check: Proportional Odds (Brant test)

The core assumption of ordinal regression is proportional odds, the coefficients that desribe relationship between categories are same. 

```{r message=FALSE}
cat("=== Brant test ===\n")
poTest(ordinal_A)
poTest(ordinal_B)
```

H0: parallel regression assumption holds. Rejection (p < 0.05) means that predictor's effect differs across categories. As we can see in the output, most of the predictors are violating this assumption, making the modeled probabilities inccorect for categories. 

# Partial Proportional Odds
As the ordinal regression did not satisfy PO assumption, it is possible to use partial proportional odds model which relaxes this assumption on predictors that violate it. The coefficients for predictors like this are fitted for each category separated while the ones that satisfy the PO will keep it same accross categories.

```{r}
library(VGAM)
fit_partial <- function(train_data, partition) {
  train_subset_partial <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  vglm(AQI_p24 ~ ., data = train_subset_partial,
       family = cumulative(parallel = TRUE ~ -1 + AQI_t2 + PM10))
}

partial_A <- fit_partial(train_data, PARTITION_A)
#partial_B <- fit_partial(train_data, PARTITION_B)
summary(partial_A)
```
In partition B, there is a much bigger number of predictors than in A. As only small amount of these do not violate PO assumption, this model is basically very close to multinominal regression (both are actually) and we need to estimate coefficients for every predictor for every category. That requires that every combination can be assigned some probability. Problem arises if the estimated probabilities are close to 0 or 1 respectively, as the link functions are based on logharitms, making the estimation unstable.   

```{r}
predict_partial <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(test_data$AQI_p24)
  probs  <- predictvglm(model, newdata = test_features, type = "response")
  preds  <- max.col(probs)
  list(actual = actual, preds = preds)
}

eval_partial <- function(model, test_data, partition) {
  p <- predict_partial(model, test_data, partition)
  compute_metrics(p$preds, p$actual)
}

class_recall_partial <- function(model, test_data, partition) {
  p <- predict_partial(model, test_data, partition)
  compute_class_recall(p$preds, p$actual)
}

list(
  "A - pollutants only" = eval_partial(partial_A, test_data, PARTITION_A)
) %>%
  list_rbind(names_to = "Partition")

cat("--- Partition A ---\n")
class_recall_partial(partial_A, test_data, PARTITION_A) %>% print()
```

### Multinomial regression
The partial proportional odds model is very close to the full multinomial regression as only two features are fulfilling the PO assumption. The next step is to try to fit full multinomial regression, even though the improvement will be very small if at all.
```{r}
MULTINOMIAL_WEIGHT_POWER <- 0.2 # 0 = no reweighting, 1 = fully balanced
compute_class_weights_multinomial <- function(train_data, power = MULTINOMIAL_WEIGHT_POWER) {
  class_counts <- table(train_data$AQI_p24)
  w_raw     <- (1 / as.numeric(class_counts))^power
  w_per_row <- w_raw[as.integer(as.character(train_data$AQI_p24))]
  # Normalize so weights sum to n (preserves effective sample size)
  w_per_row / sum(w_per_row) * nrow(train_data)
}

fit_multinom <- function(train_data, partition) {
  train_subset_multinom <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  train_subset_multinom$AQI_p24 <- as.factor(as.character(train_subset_multinom$AQI_p24))
  vglm(AQI_p24 ~ ., data = train_subset_multinom,
            family = multinomial(refLevel = 1),
            model = TRUE,
            weights = compute_class_weights_multinomial(train_subset_multinom))
}


multinomial_A <- fit_multinom(train_data, PARTITION_A)
summary(multinomial_A)
multinomial_B <- fit_multinom(train_data, PARTITION_B)
summary(multinomial_B)
exp(coef(multinomial_B))
```

We can interpret the exponentiated coefficients as odds ration against the base class, which is AQI level 1 (Good). All of the pollutant values have higher coefficients with higher category, meaning that increasing value of pollutants are more likely to equal to higher levels of AQI, as it should. For example with increasing pollutant PM2.5, it is 1.04 more likely AQI 2 than AQI 1. For third level it is 1.07 and for fourth 1.09. Values lower than 0 means that it is less likely, rain the day before usually does not mean higher level of AQI. We can visualize this by plotting proportion of rain for each category. Similar phenomena can explain large coefficients for Snowfall and Soil Moist Tempeture, but as the values are more rare it makes much bigger impact. 

```{r}
train_data %>%
  mutate(Rain_Status = ifelse(Rain > 0, "Raining", "No Rain")) %>%
  ggplot(aes(x = AQI_p24, fill = Rain_Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Rain by AQI Level",
        x = element_blank(), y = element_blank(),
        fill = "Condition") +
  theme_minimal()
```

#### Eval multinominal
The accuracy and recall for this model is not very significantly changed compared to the other parametric or non-parametric models.
```{r}
predict_multinom <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(as.character(test_data$AQI_p24))
  probs  <- predictvglm(model, newdata = test_features, type = "response")
  preds  <- max.col(probs)
  list(actual = actual, preds = preds)
}

eval_multinom <- function(model, test_data, partition) {
  p <- predict_multinom(model, test_data, partition)
  compute_metrics(p$preds, p$actual)
}

class_recall_multinom <- function(model, test_data, partition) {
  p <- predict_multinom(model, test_data, partition)
  compute_class_recall(p$preds, p$actual)
}

list(
  "A - pollutants only" = eval_multinom(multinomial_A, test_data, PARTITION_A),
  "B - all features" = eval_multinom(multinomial_B, test_data, PARTITION_B)
) %>%
  list_rbind(names_to = "Partition")

cat("--- Partition A ---\n")
class_recall_multinom(multinomial_A, test_data, PARTITION_A) %>% print()
cat("\n--- Partition B ---\n")
class_recall_multinom(multinomial_B, test_data, PARTITION_B) %>% print()
```

#### Confusion matrix
```{r}
preds_multinom <- predict_multinom(multinomial_B, test_data, PARTITION_B)
actual_cls  <- preds_multinom$actual
pred_cls    <- preds_multinom$preds
act_pred_tb <- tibble(Actual = actual_cls, Predicted = pred_cls)

plot_confusion_matrix(actual_cls, pred_cls, "Confusion Matrix: Multinomial Regression Partition B")
```

#### Assumption check: Linearity of log-odds

The multinomial logistic regression assumes that the log-odds of each class vs the reference class (Good) is a linear function of the continuous predictors. This means that for each predictor, the relationship between the predictor and the log-odds of being in a particular AQI category (Fair, Moderate, Poor) vs being in the Good category should be linear.

We check all continuous predictors in `PARTITION_B` for this assumption. We exclude some variables that are not suitable for this check, such as `Rain` and `Snowfall` which are heavily zero-inflated, `Weather_Code` which is categorical, and `hour` and `month` which are cyclic.

```{r}
skip_vars <- c("Rain", "Snowfall", "Weather_Code", "hour", "month")
continuous_vars_check <- setdiff(PARTITION_B, skip_vars)

logodds_df <- map_dfr(continuous_vars_check, function(var) {
  df <- train_data %>%
    select(x = all_of(var), y = AQI_p24) %>%
    mutate(y = as.integer(y))

  breaks <- unique(quantile(df$x, probs = seq(0, 1, length.out = 13)))

  if (length(breaks) < 3) return(NULL)
  df <- df %>% mutate(bin = cut(x, breaks = breaks, include.lowest = TRUE, labels = FALSE))

  bin_stats <- df %>%
    count(bin, y) %>%
    #complete(bin, y = 1:4, fill = list(n = 0)) %>%
    mutate(total = sum(n), prop = n / total, .by=bin) %>% # calculate proportion of each class in the bin

  bin_medians <- df %>%
    summarise(x_med = median(x), .by = bin) # calculate median of each bin

  bin_stats %>%
    left_join(bin_medians, by = "bin") %>%
    group_by(bin) %>%
    mutate(
      p_ref = prop[y == 1], # reference class proportion (Good=1) for the bin
      log_odds = log((prop) / (p_ref)) # log-odds vs reference class (Good=1)
    ) %>%
    filter(y != 1) %>%
    mutate(Class = factor(y, levels = 2:4, labels = c("Fair", "Moderate", "Poor")),
      variable = var
    )
})

ggplot(logodds_df, aes(x_med, log_odds, color = Class)) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Linearity of log-odds assumption check",
    subtitle = "log(P(AQI=k) / P(AQI=Good)) vs predictor bin median",
    x = "Predictor bin median", y = "log(P(k) / P(Good))", color = "AQI Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom"
  )
```

The results are somewhat mixed. For some predictors, the assumptions holds, but is violated for others. AQI_tX features show a very strong linear relationship with log-odds (which makes sense given the temporal autocorrelation of AQI). However, for some weather features (e.g. `O3`, `Wind_Direction`) the relationship is clearly non-linear, which violates the assumption and may lead to suboptimal performance of the multinomial model compared to more flexible methods like Random Forest or XGBoost that do not rely on this assumption.

## Models Evaluation Summary

```{r}
# Collect test-set metrics for every fitted model into one tibble
summary_table <- bind_rows(
  eval_rf(rf_A, test_data, PARTITION_A)            %>% mutate(Method = "Random Forest",      Type = "Non-parametric", Partition = "A"),
  eval_rf(rf_B, test_data, PARTITION_B)            %>% mutate(Method = "Random Forest",      Type = "Non-parametric", Partition = "B"),
  eval_xgb(xgb_A, dtest_A, test_data)              %>% mutate(Method = "XGBoost",            Type = "Non-parametric", Partition = "A"),
  eval_xgb(xgb_B, dtest_B, test_data)              %>% mutate(Method = "XGBoost",            Type = "Non-parametric", Partition = "B"),
  eval_knn(knn_A, test_data)                       %>% mutate(Method = "kNN",                Type = "Non-parametric", Partition = "A"),
  eval_knn(knn_B, test_data)                       %>% mutate(Method = "kNN",                Type = "Non-parametric", Partition = "B"),
  eval_ordinal(ordinal_A, test_data, PARTITION_A)  %>% mutate(Method = "Ordinal Regression", Type = "Parametric",     Partition = "A"),
  eval_ordinal(ordinal_B, test_data, PARTITION_B)  %>% mutate(Method = "Ordinal Regression", Type = "Parametric",     Partition = "B"),
  eval_partial(partial_A, test_data, PARTITION_A)  %>% mutate(Method = "Partial PPO",        Type = "Parametric",     Partition = "A"),
  eval_multinom(multinomial_A, test_data, PARTITION_A) %>% mutate(Method = "Multinomial Reg.", Type = "Parametric",   Partition = "A"),
  eval_multinom(multinomial_B, test_data, PARTITION_B) %>% mutate(Method = "Multinomial Reg.", Type = "Parametric",   Partition = "B")
) %>%
  select(Type, Method, Partition, `Acc (%)`, `+-1 Class Acc (%)`, `Macro F1`, `Poor Recall (%)`) %>%
  arrange(Type, Method, Partition)

summary_table
```

**Scenario 1 - Models across 2 feature space partitions**
Across all three models families (Random Forest, XGBoost, kNN), adding weather features (Partition B) consistently raises Macro F1 and accuracy over pollutants-only (Partition A). This shows that weather variables carry additional predictive signal regardless of the model choice. 

@TODO: add feature importance plots, compare them


`Features that matter in both families (e.g. `Pressure_MSL`, lagged AQI, `O3`) are therefore the most robust candidate features for this task. Non-parametric models expose this through permutation importance; parametric models through coefficient size and significance, offering complementary angles on explainability.`

**Scenario 2 - Parametric vs. Non-parametric models**
@TODO: dopisat
`Non-parametric methods (RF, XGBoost) achieve the highest Macro F1 scores because they capture non-linear interactions (e.g. the non-monotone O3-AQI relationship shown in the EDA) without any distributional assumption. Parametric methods (Ordinal Regression, Partial PPO, Multinomial Regression) score lower on F1 but offer stronger explainability. Coefficients have a direct probabilistic interpretation, standard errors and p-values are available, and the Brant test makes the proportional-odds assumption auditable. They are also fully reproducible, as the same data always produce the same model, with no randomness from tree bootstrapping or gradient steps. For a regulatory or public-health context where decisions must be justified, parametric models are preferable; for raw predictive power, non-parametric ensemble methods win.`

## Scenario 3 - Feature Selection: Wrapper vs. Embedded

The multinomial model fitted above uses all 22 features in PARTITION_B. By utilizing feature selection methods, it may be possible to find smaller set of predictors that would explain the data better than the full 22 features.
Two embedded methods and one wrapper one was performed and compared between each other and the base multionomial model.

### Parameters

```{r}
library(glmnet)

LASSO_ALPHA <- 1.0 # pure L1 penalty (Lasso)
ELNET_ALPHA <- 0.5 # 50/50 L1+L2 blend (Elastic Net)
GLMNET_NFOLDS <- 5 # folds for cross-validated lambda selection
```

### Lasso

Lambda is selected by 10-fold cross-validation. Both the min lambda and 1-SE lambda are computed and showcase on the plot. For evaulation, we will use 1-SE lambda as it picks simpler model within one standard error of minimum. The plotted lambdas showcases that the 1-SE lambda keeps 17 features in the model, while minimal does 20. This does not mean the model will only retain 17 predictors because as we do multinomial regression and estimating coefficent for each class separately, some predictors can be included in one class estimation and not in others.

```{r}
set.seed(2026)
cv_lasso <- cv.glmnet(
  as.matrix(train_data %>% select(all_of(PARTITION_B))),
  train_data$AQI_p24,
  family   = "multinomial",
  alpha    = LASSO_ALPHA,
  nfolds   = GLMNET_NFOLDS,
  weights  = compute_class_weights_multinomial(train_data),
  type.measure = "class"
)
plot(cv_lasso)
title("Lasso - CV misclassification vs. log(lambda)", line = 2.5)
cat("Lasso lambda.min:", cv_lasso$lambda.min, "\n")
cat("Lasso lambda.1se:", cv_lasso$lambda.1se, "\n")
coef(cv_lasso)
```

### Elastic Net

Elastic Net adds an L2 penalty alongside L1. The L2 term encourages correlated features to receive similar (non-zero) coefficients rather than removing one of them.

```{r}
set.seed(2026)
cv_elnet <- cv.glmnet(
  as.matrix(train_data %>% select(all_of(PARTITION_B))),
  train_data$AQI_p24,
  family   = "multinomial",
  alpha    = ELNET_ALPHA,
  nfolds   = GLMNET_NFOLDS,
  weights  = compute_class_weights_multinomial(train_data),
  type.measure = "class"
)
plot(cv_elnet)
title("Elastic Net - CV misclassification vs. log(lambda)", line = 2.5)
cat("Elastic Net lambda.min:", cv_elnet$lambda.min, "\n")
cat("Elastic Net lambda.1se:", cv_elnet$lambda.1se, "\n")
```

### Backward step

Starting from the full multinomial B model, features are removed one at a time. At each step the feature whose removal most reduces AIC is dropped, continuing until no removal improves AIC. In this case removing any features lead to worse model so all of the predictors are kept.

```{r}
multinom_step_B <- step4vglm(multinomial_B, direction = "backward")
cat("Full model AIC:    ", AIC(multinomial_B),   "\n")
cat("Reduced model AIC: ", AIC(multinom_step_B), "\n")
cat("Features retained:", length(attr(terms(multinom_step_B), "term.labels")), "/", length(PARTITION_B), "\n")
```

### Feature retention table
Summary table that showcase all the features and if they were kept across feature selection methods. In the case of this dataset and predicting AQI, all of the features were kept in the models, except in pure Lasso regression, where variable Weather_Code was removed.

```{r}
extract_kept_features <- function(cv_model, s = "lambda.1se") {
  coef_list  <- coef(cv_model, s = s)
  result_list <- lapply(names(coef_list), function(cl) {
    mat <- as.matrix(coef_list[[cl]])
    tibble(class = cl, feature = rownames(mat), coef = as.numeric(mat))
  })
  bind_rows(result_list) %>%
    group_by(feature) %>%
    summarise(kept = any(coef != 0), .groups = "drop") %>%
    filter(feature != "(Intercept)")
}

kept_lasso <- extract_kept_features(cv_lasso) %>% rename(Lasso_1se = kept)
kept_elnet <- extract_kept_features(cv_elnet) %>% rename(ElNet_1se = kept)

feature_selection_table <- kept_lasso %>%
  left_join(kept_elnet, by = "feature") %>%
  mutate(
    StepAIC = feature %in% attr(terms(multinom_step_B), "term.labels"),
    # Ensure NAs (if any) are treated as FALSE
    Lasso_1se = replace_na(Lasso_1se, FALSE),
    ElNet_1se = replace_na(ElNet_1se, FALSE),
    StepAIC   = replace_na(StepAIC, FALSE),
    # Count how many methods selected the feature
    Kept = as.integer(Lasso_1se) + as.integer(ElNet_1se) + as.integer(StepAIC)
  ) %>%
  arrange(desc(Kept), feature)
print(feature_selection_table, n=22)
```

### Performance comparison
Comparing performances, we can see that Lasso and Elasticnet methods perform worse on most of the metrics. The stepAIC backward model did keep all of the features, making it identical to the default multinomial model.

```{r}
test_mat <- as.matrix(test_data %>% select(all_of(PARTITION_B)))
preds_lasso  <- as.integer(predict(cv_lasso, newx = test_mat, s = "lambda.1se", type = "class"))
preds_elnet  <- as.integer(predict(cv_elnet, newx = test_mat, s = "lambda.1se", type = "class"))
preds_step   <- predict_multinom(multinom_step_B, test_data, PARTITION_B)$preds
preds_multinom_B <- predict_multinom(multinomial_B, test_data, PARTITION_B)$preds
actual       <- as.integer(as.character(test_data$AQI_p24))

feature_selection_metrics <- bind_rows(
  compute_metrics(preds_multinom_B, actual) %>% mutate(Method = "Multinomial B (all)"),
  compute_metrics(preds_step,   actual) %>% mutate(Method = "StepAIC backward (all)"),
  compute_metrics(preds_lasso,  actual) %>% mutate(Method = "Lasso lambda.1se (- Weather_Code)"),
  compute_metrics(preds_elnet,  actual) %>% mutate(Method = "Elastic Net lambda.1se (all)")
) %>%
  select(Method, everything())

feature_selection_metrics

```

#### Per-class recall
The biggest difference is that the penalized models very poorly classified the Good category compared to the default multinomial model. It performed worst on the Poor category too, the only improvement can be seen in Fair one which does not provide any meaningful value.
```{r}
feature_selection_recall <- bind_rows(
  compute_class_recall(preds_multinom_B, actual) %>% mutate(Method = "Multinomial B"),
  compute_class_recall(preds_step,   actual) %>% mutate(Method = "StepAIC"),
  compute_class_recall(preds_lasso,  actual) %>% mutate(Method = "Lasso"),
  compute_class_recall(preds_elnet,  actual) %>% mutate(Method = "Elastic Net")
) %>%
mutate(Label = factor(Label, levels = c("Good", "Fair", "Moderate", "Poor")))
feature_selection_recall

feature_selection_recall %>%
ggplot(aes(x = Label, y = `Recall (%)`, fill = Method)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Per-class recall — Scenario 3 methods and full multinomial",
    x = element_blank(), y = "Recall (%)", fill = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Interpretation

The Lasso method was the only method that removed predictor, in this case it was Weather_Code variable. The one possible reason could be that the variable is already explained by other weather conditions like rain, snowfall or cloud coverage and hence is dropped by Lasso regression. The Elastic Net method kept all the features.

The wrapper method, backward step selection, did also keep all the variables, making the model same as the multinomial B model. All of these methods suggest that the picked predictors are useful and provide new information about AQI.

The performance was not increased in any meaningful way. The step model was in the end same as the base multinomial model and the Lasso and Elastic Net performed worse, mainly on Good class but all the others too, except the Fair class which does not add any meaningful value.

The feature set should be kept as it is, with all the features present in each model. The preliminary feature selection was based on domain knowledge and filter methods like VIF. The embedded and wrapper methods did not showcase any major improvement, just confirmed that predictor subset was picked correctly and every variable does provide meaningful information to the prediction.

## Scenario 4 - Tree And Heatmap Visualizations (vs. Models)

### Decision Tree

Our goal here is to fit a single decision tree on the PARTITION_B features used by the RF model, then ask:

1. Which features does the tree actually use?
2. What rules does it learn?
3. At what depth does a single tree "agree" with the RF? (as much as possible)
4. What does the tree fail to do that the RF succeeds at, and why?

#### Parameters and fit

```{r}
library(rpart)
library(rpart.plot)

TREE_MAXDEPTH  <- 5 # max depth of the tree
TREE_CP <- 0.001 # complexity parameter: min relative improvement per split
TREE_MINSPLIT <- 20 # min obs in a node to attempt a split
TREE_MINBUCKET <- 7 # min obs in any terminal leaf

# RF weighted predictions
preds_rf_B <- preds_B$preds
actual_int  <- preds_B$actual
```

Why these values?

- `maxdepth = 5` balances readability against complexity;
  the depth sweep below shows the tree plateaus at depth 8 anyway.
- `cp = 0.001` is deliberately small — we want the tree to grow as far as it profitably can,
  so we can see where it naturally stops.
- `minsplit = 20, minbucket = 7` prevent overfitting to tiny node populations.

```{r}
tree_B <- rpart(
  AQI_p24 ~ .,
  data    = train_data %>% select(all_of(c(PARTITION_B, "AQI_p24"))),
  method  = "class",
  control = rpart.control(
    maxdepth  = TREE_MAXDEPTH,
    cp        = TREE_CP,
    minsplit  = TREE_MINSPLIT,
    minbucket = TREE_MINBUCKET
  )
)

preds_tree_B <- as.integer(predict(tree_B, newdata = test_data %>% select(all_of(PARTITION_B)), type = "class"))
```

#### Complexity table

The table shows the cross-validation estimates of misclassification error (`xerror`) at each split. We try to find the optimal tree by looking for the point where `xerror` stops decreasing (beyond that, generalisation does not improve).

```{r}
printcp(tree_B)
```

The `xerror` stops decreasing at around 13 splits, tells us that a single tree on this
problem exhausts its useful complexity at that number of leaves. More splits would likely only lead to overfitting.

#### Which features does the tree use?

```{r}
vars_used <- unique(tree_B$frame$var[tree_B$frame$var != "<leaf>"])
vars_not_used <- setdiff(PARTITION_B, vars_used)

tibble(
  Status   = c(rep("Used", length(vars_used)), rep("Not used", length(vars_not_used))),
  Feature  = c(vars_used, vars_not_used)
) %>% print(n = Inf)
```

- AQI_t - This is the root split, which confirms the strong lag-1 autocorrelation from our EDA
- AQI_t1, AQI_t2, AQI_t24 - These lags are not used, likely because they are highly correlated with AQI_t and do not provide additional information for splits.
- Pressure_MSL again appears just like in other models, where it was explained
- Dew_Point, PM25, O3, Soil_Moisture_0_7, Wind_Speed - these appear at deeper nodes.
- Cloud cover variables, Rain, Snowfall, Weather_Code, hour, NO2, SO2, Wind_Direction - they either carry little additional information, or that signal is already captured by other features

#### How deep should the tree be?

We sweep `maxdepth` from 2 to 12 and evaluate the following:
1. agreement with the weighted RF on the test set,
2. overall test accuracy

```{r}
depth_sweep <- tibble(maxdepth = 2:12) %>%
  mutate(
    fit = map(maxdepth, ~ rpart(
      AQI_p24 ~ .,
      data    = train_data %>% select(all_of(c(PARTITION_B, "AQI_p24"))),
      method  = "class",
      control = rpart.control(
        maxdepth = .x, cp = TREE_CP,
        minsplit = TREE_MINSPLIT, minbucket = TREE_MINBUCKET
      ) # each depth gets its own fit
    )),
    preds_tree = map(fit, ~ as.integer(
      predict(.x, newdata = test_data %>% select(all_of(PARTITION_B)), type = "class")
    )), # each fit gets its own preds
    agreement_rf_weighted = map_dbl(preds_tree, ~ mean(.x == preds_rf_B)), 
    accuracy = map_dbl(preds_tree, ~ mean(.x == actual_int)),
    n_leaves = map_int(fit, ~ sum(.x$frame$var == "<leaf>"))
  ) %>%
  select(-fit, -preds_tree)

print(depth_sweep)
```

```{r}
depth_sweep %>%
  pivot_longer(c(agreement_rf_weighted, accuracy), names_to = "metric", values_to = "value") %>%
  mutate(metric = case_when(
    metric == "agreement_rf_weighted" ~ "Agreement with RF (weighted)",
    metric == "accuracy" ~ "Test accuracy (vs true labels)"
  )) %>%
  ggplot(aes(maxdepth, value * 100, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_text(
    data = depth_sweep %>%
      group_by(maxdepth) %>%
      summarise(n_leaves = first(n_leaves),
                y = max(agreement_rf_weighted, accuracy) * 100 + 1.5),
    aes(maxdepth, y, label = paste0(n_leaves, "L")),
    color = "grey40", size = 3
  ) +
  scale_x_continuous(breaks = 2:12) +
  scale_color_manual(values = c(
    "Agreement with RF (weighted)"  = "#2c7bb6",
    "Test accuracy (vs true labels)" = "#d7191c"
  )) +
  labs(
    title = "Decision tree depth vs. RF agreement and test accuracy",
    x = "Tree max depth", y = "(%)", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Both curves plateau at depth = 8 with 23 leaves. Deeper growth adds no leaves (no split improves training error by cp)
RF Agreement ceiling is about 78%. The rest is predictive advantage of combination of the 500 trees (ensemble resolves ambiguity that a single tree cannot).
Single tree has accuracy of 59.7% vs. unweighted RF 61.9% - it is simpler and more interpretable, but the RF has better accuracy.

For interpretation, we stick with depth = 5, which is more readable.

#### Visualisation 

```{r}
rpart.plot(
  tree_B,
  type = 4, # display split labels at all nodes
  extra = 104, # show all-class probabilities + leaf percentage
  tweak = 1.2,
  box.palette = list("#50ccaa", "#f0e641", "#ff5050"),
  main = paste0("Decision Tree - AQI 24h-ahead prediction (depth = ", TREE_MAXDEPTH,", balanced)")
)
```

At the root, the tree splits on `AQI_t < 3`, which is the Moderate threshold. When `AQI_t` is below 3, it looks into `PM25` value. If `PM25` is below 8.5 µg/m³, the tree predicts Fair mostly, except in late autumn and winter (month >= 11) where lower wind speed and higher soil moisture tip a node toward Moderate. Above 8.5 µg/m³, atmospheric pressure becomes decisive - low-pressure conditions favour dispersal and the tree predicts Fair, while a high-pressure lid traps pollutants and shifts the prediction to Moderate, with dew point providing a seasonal refinement. 
In the right branch, pressure again divides the tree: below 1018 hPa, O3 and dew point together determine whether the outcome is Fair or Moderate (along with another check for pressure). Above 1018 hPa, the `AQI_t` level provides the decisive second split, and only at the extreme — `AQI_t >= 4` combined with very high pressure (>= 1030 hPa) - does the tree reach a leaf predicting Poor, with 77% confidence. The tree never predicts Good, as there are very few training observations in that class.

#### Class imbalance and its effect on tree predictions

```{r}
tibble(
  Class = aqi_levels,
  Predicted = as.integer(table(factor(preds_tree_B, levels = 1:4))),
  Actual = as.integer(table(factor(actual_int, levels = 1:4)))
) %>% print()
```

Currently, the tree we tested predicts *Fair* and *Moderate* most of the time, and never predicts *Good* and only rarely predicts *Poor* (205 times), even though those classes together make up around 2000 rows in the test set. We can address this by setting class priors to be uniform at the cost of overall accuracy.

#### Balanced tree

We try to address this by setting uniform class priors (weights) which tells rpart to treat all classes as equally likely before seeing any data, effectively upweighting minority classes. We didn't use the inverse weights as in the RF, because in this case (single tree) it would be too extreme.

```{r}
tree_B_balanced <- rpart(
  AQI_p24 ~ .,
  data = train_data %>% select(all_of(c(PARTITION_B, "AQI_p24"))),
  method = "class",
  parms = list(prior = c(0.25, 0.25, 0.25, 0.25)),  # equal priors
  control = rpart.control(
    maxdepth = TREE_MAXDEPTH,
    cp = TREE_CP,
    minsplit = TREE_MINSPLIT,
    minbucket = TREE_MINBUCKET
  )
)

preds_tree_bal <- as.integer(predict(tree_B_balanced,
  newdata = test_data %>% select(all_of(PARTITION_B)), type = "class"))
```

#### Visualisation (balanced tree)

```{r}
rpart.plot(
  tree_B_balanced,
  type = 4, # display split labels at all nodes
  extra = 104, # show all-class probabilities + leaf percentage
  tweak = 1.2,
  box.palette = list("#50f0e6", "#50ccaa", "#f0e641", "#ff5050"),
  main = paste0("Decision Tree - AQI 24h-ahead prediction (depth = ", TREE_MAXDEPTH,", balanced)")
)
```

With the class weights, the first split still falls on `AQI_t`, with the treshold staying the same. On the left branch, the `AQI_t` decides the next split again - if it's less than 2, the tree immediately predicts *Good* with 75% probability. For `AQI_t` of 2 and higher, the tree splits on `Pressure_MSL` - whether its below 1016 hPa, then decides based on the `month` and `SO2` values. On the right branch of the pressure split, where the pressure is above or equal to 1016 hPa, the tree splits on `O3` - if it's below 71 µg/m³, the tree further splits on `NO2` to decide between *Fair* and *Good*, while if it's above that threshold, it predicts *Poor* if `O3` is above 87 µg/m³, otherwise it predicts *Fair*.

On the right branch of the root, the tree first splits on `AQI_t` again, with a threshold of 4. If it is above or equal to 4, it always predicts *Poor* with 77% confidence, while if it is below 4, it splits on `O3` - if `O3` is below 100 µg/m³, it decides based on pressure, dew point and PM25. If `O3` is above 100 µg/m³, it predicts *Poor* with 60% confidence.

```{r}
tibble(
  Class = 1:4,
  Label = aqi_levels,
  Actual_n = as.integer(table(actual_int))
) %>%
  mutate(
    Tree_unweighted = sapply(Class, function(k)
      round(100 * sum(preds_tree_B == k & actual_int == k) / sum(actual_int == k), 1)),
    Tree_balanced = sapply(Class, function(k)
      round(100 * sum(preds_tree_bal == k & actual_int == k) / sum(actual_int == k), 1)),
    RF_weighted = sapply(Class, function(k)
      round(100 * sum(preds_rf_B == k & actual_int == k) / sum(actual_int == k), 1))
  ) %>% print()


tibble(
  Model = c("Tree (unweighted)", "Tree (balanced)", "RF (weighted)"),
  Accuracy = round(c(
    mean(preds_tree_B == actual_int),
    mean(preds_tree_bal == actual_int),
    mean(preds_rf_B == actual_int)
  ) * 100, 1),
  Macro_F1 = round(c(
    macro_f1(preds_tree_B, actual_int),
    macro_f1(preds_tree_bal, actual_int),
    macro_f1(preds_rf_B, actual_int)
  ), 3)
) %>% print()
```

```{r}
recall_comparison %>%
  pivot_longer(c(Tree_unweighted, Tree_balanced, RF_weighted), names_to = "Model", values_to = "Recall") %>%
  mutate(
    Model = case_when(
      Model == "Tree_unweighted" ~ "Tree (unweighted)",
      Model == "Tree_balanced" ~ "Tree (balanced priors)",
      Model == "RF_weighted" ~ "RF (weighted)"
    ),
    Model = factor(Model, levels = c(
      "Tree (unweighted)", "Tree (balanced priors)", "RF (weighted)")),
    Label = factor(Label, levels = aqi_levels)
  ) %>%
  ggplot(aes(Label, Recall, fill = Model)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  scale_fill_manual(values = c(
    "Tree (unweighted)" = "#d7191c",
    "Tree (balanced priors)" = "#fdae61",
    "RF (weighted)" = "#2c7bb6"
  )) +
  labs(
    title    = "Per-class recall: single tree vs. weighted random forest",
    subtitle = "Unweighted tree collapses minority classes. Balanced priors recover recall but sacrifice accuracy.",
    x = "AQI class", y = "Recall (%)", fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Summary and evaluation**

| Model | Good recall | Poor recall | Overall acc | Macro F1 | Interpretability |
|---|---|---|---|---|---|
| Tree (unweighted) | 0% | 1.6% | 59.4% | 0.313 | 14 leaves, 8 features |
| Tree (balanced) | 75.7% | 59.5% | 36.4% | 0.339 | more leaves, same features |
| RF (weighted) | 23.2% | 53.3% | 58.7% | **0.464** | 500 trees, black-box |

- The unweighted tree is the most interpretable, but ignores the minority classes completely. It is useful for describing the majority patterns in the training data, but would fail badly in an early-warning use-case where catching *Poor* air quality is critical. 

- The balanced tree recovers minority-class recall dramatically (Good: 0%->76%, Poor: 2%->60%) at the cost of heavy over-prediction of those classes (predicts Good 6742 times vs. 1402 actual).

- The weighted RF achieves the best *macro F1* (0.464) by combining 500 trees with inverse-frequency weights, recovering minority-class recall without totally sacrificing majority-class precision.

A single tree of any depth agrees with the RF on at most ~78% of test cases.  The remaining 22% represent ambiguous observations where the ensemble finds signal by averaging over many weak learners, while a single rule-based partition cannot. This is a fundamental limit of the single-tree model class on this dataset.

# References
- [1] [Air Quality in Slovakia - Ministry of Health](https://www.minzp.sk/files/oblasti/ovzdusie/ochrana-ovzdusia/dokumenty/strategia-ochrany-ovzdusia/hodnotenie_pre_strategiu_podklady.pdf)
- [2] [European Air Quality Index](https://www.eea.europa.eu/en/analysis/maps-and-charts/index)