---
title: "AQI Prediction - Elements of AI Capstone Project"
author: "Filip Škrabák & Patrik Prizbul"
date: "2026-03-15"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(arrow)
library(lubridate)
library(tsibble)
library(here) # for handling paths correctly
library(corrplot)
library(zoo) # for na.approx
library(forecast)
library(generics)
library(fable)
library(feasts)
library(ggtime) ## gg_season
here::i_am("project.Rmd")
```

# Introduction
The main objective of this project is to predict the Air Quality Index (AQI) for Bratislava, Slovakia, using historical air quality and weather data. This requires us to merge two datasets - the pollutants dataset and the weather dataset.

# Pollutants Dataset

-   This dataset comes from [European Enviroment Agency (EEA) Air Quality Download Service](https://eeadmz1-downloads-webapp.azurewebsites.net/)
-   It contains pollutants from year 2017 until the end of 2024
-   Dataset type is Primary Validated Data (E1a), which is verified
-   Timezone is stated as UTC+1

First, we load the dataset and take a look at its structure.

```{r}
pollutants_data <- open_dataset(here("pollutants"))

glimpse(pollutants_data)
```

There are 12 columns and 902 260 rows. First, lets keep only the concerned station (SK0001A), the years of interest (2017-2024) and relevant columns.

```{r}
filtered_air_data <- pollutants_data %>%
  select(Start, Pollutant, Value, Samplingpoint, Validity) %>% # lazy loading
  filter(grepl("SK0001A", Samplingpoint)) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024) %>%
  collect()
```

Lets look at numeric ranges and unique values of the columns.

```{r}
summary(filtered_air_data)
```

The summary suggests that:
- The `Value` column contains some negative values, which are probably sensor errors.
- The `Validity` is either 1 or 2. According to data source, both are valid, with 2 being below detection limit. We are keeping both.

Next, we should translate the `Pollutant` codes into human readable names and reshape the data to have columns for each pollutant. Finally, we make sure the data is sorted by measurement time and we convert it into a tsibble for time series analysis.

```{r}
clean_air_data <- filtered_air_data %>%
  # Translate EU codes into factors
  mutate(Pollutant_Name = case_when(
    Pollutant == 1 ~ "SO2",
    Pollutant == 5 ~ "PM10",
    Pollutant == 7 ~ "O3",
    Pollutant == 8 ~ "NO2",
    #Pollutant == 10 ~ "CO",
    Pollutant == 6001 ~ "PM2.5",
    TRUE ~ as.character(Pollutant)
  )) %>%
  select(Start, Pollutant_Name, Value) %>%
  pivot_wider(
    names_from = Pollutant_Name, 
    values_from = Value
  ) %>%
  arrange(Start) %>%
  as_tsibble(index = Start)

glimpse(clean_air_data)

View(clean_air_data)
```

Now we have a cleaned dataset with 5 pollutants. Lets check whether our time series has any gaps. Our data is hourly.

Lets check gaps in our timeseries whether there are any missing hours. For now, we will 

```{r}
gaps <- clean_air_data %>%
  scan_gaps()

nrow(gaps)
gaps
```

There are 410 gaps. For now, lets fill them with NA values. Also check the tsibble gaps again to make sure they are filled.
```{r}
clean_air_data <- clean_air_data %>%
  fill_gaps()

has_gaps(clean_air_data)

interval(clean_air_data)
duplicates(clean_air_data)
```

The output confirms there are no more gaps, the interval is 1 hour and there are no duplicates.

```{r}
complete_rows <- sum(complete.cases(clean_air_data)) # 60845
rows <- nrow(clean_air_data) # 70128

missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage

# longest gap
clean_air_data %>%
  as_tibble() %>%
  pivot_longer(-Start, names_to = "pollutant", values_to = "value") %>%
  arrange(pollutant, Start) %>%
  group_by(pollutant) %>%
  mutate(is_na = is.na(value),
         group = cumsum(!is_na & !lag(!is_na, default = TRUE))) %>%
  filter(is_na) %>%
  group_by(pollutant, group) %>%
  summarise(gap_start = min(Start), gap_end = max(Start), hours = n(), .groups = "drop") %>%
  arrange(desc(hours)) %>%
  select(-group) %>%
  mutate(days = hours / 24)

```

13.24% of our data is missing. The longest gap (sequence of NA values) for any pollutant is 10.2 days for NO2.

Next step is to impute the missing values. We use a 2 step approach. 
- Linear interpolation for small gaps (smaller than 8 hours)
- Mean for same hour of day - for larger gaps, this respects the cycles in the data

```{r}
impute_missing_values <- function(x, timestamps, max_linear = 8, window_days = 3) {
  x_out <- zoo::na.approx(x, na.rm = FALSE, maxgap = max_linear)

  hrs <- lubridate::hour(timestamps)
  n   <- length(x_out)

  global_hour_mean <- tapply(x, hrs, mean, na.rm = TRUE)

  for (i in seq_len(n)) {
    if (!is.na(x_out[i])) next

    neighbours <- max(1, i - window_days * 24) : min(n, i + window_days * 24)

    same_hour_and_known <- neighbours[
      hrs[neighbours] == hrs[i] & !is.na(x_out[neighbours])
    ]

    x_out[i] <- if (length(same_hour_and_known) > 0) {
      mean(x_out[same_hour_and_known]) 
    } else {
      global_hour_mean[[ as.character(hrs[i]) ]] 
    }
  }
  x_out
}

complete_air_data <- clean_air_data %>%
  mutate(across(c(SO2, PM10, O3, NO2, `PM2.5`), ~ impute_missing_values(.x, Start, max_linear = 8, window_days = 3)))

complete_rows <- sum(complete.cases(complete_air_data)) # 60845
rows <- nrow(complete_air_data) # 70128
missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage
```

Now, the dataset is complete with no missing values.

# Weather Dataset

-   This dataset comes from [Open Meteo Historical Weather](https://open-meteo.com/en/docs/historical-weather-api)
-   It contains weather data from year 2017 until the end of 2024
-   Timezone is stated as GMT+0

Loading the dataset and merging it with pollutants:
```{r}
weather_file_path <- here("weather", "open-meteo-48.12N17.10E163m.csv")

# TODO: select more variables, reason about which ones
weather_data <- read_csv(weather_file_path, skip = 3) %>%
  rename(
    Start = time,
    Temperature = `temperature_2m (°C)`,
    Rain = `rain (mm)`,
    Wind_Speed = `wind_speed_10m (km/h)`
  ) %>%
  #select(Start, Temperature, Rain, Wind_Speed) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024)

complete_rows_weather <- sum(complete.cases(weather_data))
rows_weather <- nrow(weather_data)
missing_percentage_weather <- (rows_weather - complete_rows_weather) / rows_weather * 100
missing_percentage_weather

complete_data <- complete_air_data %>%
  inner_join(weather_data, by = "Start")

glimpse(complete_data)
summary(complete_data)
has_gaps(complete_data)
```

The open meteo dataset has no missing values. The merge has not changed the row count, and our time series is has no gaps. Currently, we have 36 columns.

Next step is to add AQI.

```{r}

# AQI levels and colors
aqi_levels <- c("Good", "Fair", "Moderate", "Poor", "Very poor", "Extremely poor")
aqi_colors <- c("Good"           = "#50f0e6",
                "Fair"           = "#50ccaa",
                "Moderate"       = "#f0e641",
                "Poor"           = "#ff5050",
                "Very poor"      = "#960032",
                "Extremely poor" = "#7d2181")

## European AQI
add_eaqi <- function(df) {
  df %>%
    mutate(
      AQI_PM2.5 = case_when(
        `PM2.5` <=   5 ~ 1L, `PM2.5` <=  15 ~ 2L, `PM2.5` <=  50 ~ 3L,
        `PM2.5` <=  90 ~ 4L, `PM2.5` <= 140 ~ 5L, `PM2.5` >  140 ~ 6L
      ),
      AQI_PM10 = case_when(
        PM10 <=  15 ~ 1L, PM10 <=  45 ~ 2L, PM10 <= 120 ~ 3L,
        PM10 <= 195 ~ 4L, PM10 <= 270 ~ 5L, PM10 >  270 ~ 6L
      ),
      AQI_O3 = case_when(
        O3 <=  60 ~ 1L, O3 <= 100 ~ 2L, O3 <= 120 ~ 3L,
        O3 <= 160 ~ 4L, O3 <= 180 ~ 5L, O3 >  180 ~ 6L
      ),
      AQI_NO2 = case_when(
        NO2 <=  10 ~ 1L, NO2 <=  25 ~ 2L, NO2 <=  60 ~ 3L,
        NO2 <= 100 ~ 4L, NO2 <= 150 ~ 5L, NO2 >  150 ~ 6L
      ),
      AQI_SO2 = case_when(
        SO2 <=  20 ~ 1L, SO2 <=  40 ~ 2L, SO2 <= 125 ~ 3L,
        SO2 <= 190 ~ 4L, SO2 <= 275 ~ 5L, SO2 >  275 ~ 6L
      ),
      AQI = pmax(AQI_PM2.5, AQI_PM10, AQI_O3, AQI_NO2, AQI_SO2, na.rm = TRUE),
      AQI_Label = factor(
        case_when(
          AQI == 1L ~ aqi_levels[1],
          AQI == 2L ~ aqi_levels[2],
          AQI == 3L ~ aqi_levels[3],
          AQI == 4L ~ aqi_levels[4],
          AQI == 5L ~ aqi_levels[5],
          AQI == 6L ~ aqi_levels[6],
        ),
        levels = aqi_levels,
        ordered = TRUE
      )
    )
}

complete_data <- add_eaqi(complete_data)
glimpse(complete_data)
```

# EDA

```{r}
theme_set(theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"))) 
```

## AQI Distribution

```{r}
complete_data %>%
  count(AQI_Label) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ggplot(aes(x = AQI_Label, y = n, fill = AQI_Label)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", percent)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = aqi_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "AQI Distribution (2017–2024)", x = "AQI", y = "Count")
```


## Decomposition and seasonality

STL decomposition of AQI:

```{r}
complete_data %>%
  model(STL(AQI ~ trend() + season("1 year"), robust = TRUE)) %>%
  components() %>%
  autoplot(linewidth = 0.001) +
  labs(title = "STL Decomposition of AQI", x = "Time", y = "AQI") +
  theme_minimal()

noise_season_year <- complete_data %>%
  model(STL(AQI ~ trend() + season("1 year"), robust = TRUE)) %>%
  components()

```

```{r}
monthly_aqi <- complete_data %>%
  as_tibble() %>%
  mutate(Month = yearmonth(Start)) %>%
  group_by(Month) %>%
  summarise(AQI = mean(AQI), .groups = "drop") %>%
  as_tsibble(index = Month)

monthly_aqi %>%
  gg_season(AQI, labels = "right") +
  scale_y_continuous(breaks = 1:6,
                     labels = c("Good", "Fair", "Moderate", "Poor", "Very poor", "Extremely poor"),
                     limits = c(1, 6)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017–2024)") +
  theme_minimal()
```

The seasonal plot clearly shows a **winter peak (Nov–Mar)** and **summer trough (May–Sep)** consistent across all years, driven primarily by PM2.5 from residential heating.

```{r}
hourly_aqi <- complete_data %>%
  as_tibble() %>%
  mutate(Hour = hour(Start)) %>%
  group_by(Hour) %>%
  summarise(
    Mean  = mean(AQI, na.rm = TRUE),
    Q25   = quantile(AQI, 0.25, na.rm = TRUE),
    Q75   = quantile(AQI, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hourly_aqi, aes(x = Hour)) +
  geom_ribbon(aes(ymin = Q25, ymax = Q75), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = Mean), color = "steelblue", linewidth = 1) +
  geom_point(aes(y = Mean), color = "steelblue", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(
    breaks = 1:6,
    labels = c("Good", "Fair", "Moderate", "Poor", "Very poor", "Extremely poor"),
    limits = c(1, 6)
  ) +
  labs(
    title = "Diurnal AQI pattern (hour of day, 2017–2024)",
    subtitle = "Line = mean, ribbon = IQR (25th–75th percentile)",
    x = "Hour of day", y = "AQI level"
  ) +
  theme_minimal()
```

The diurnal plot reveals two daily peaks: a **morning rush (7–9h)** and an **evening peak (18–22h)**, matching traffic and heating cycles. The overnight trough (3–5h) reflects reduced emissions and boundary layer recovery. The relatively flat mean (within ~0.3 AQI units) suggests hour-of-day is a weaker driver than season, but the wide IQR band shows high within-hour variability — winter nights pull the upper quartile sharply upward.

```{r}
complete_data %>%
  as_tibble() %>%
  mutate(
    Week = floor_date(Start, "week"),
    AQI_Label = factor(AQI_Label, levels = aqi_levels)
  ) %>%
  group_by(Week) %>%
  summarise(AQI = mean(AQI, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Week, y = AQI)) +
  geom_line(color = "steelblue", linewidth = 0.4, alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.08, se = FALSE, color = "firebrick", linewidth = 1) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    breaks = 1:6,
    labels = aqi_levels,
    limits = c(1, 6)
  ) +
  labs(
    title = "Weekly average AQI (2017–2024)",
    subtitle = "Red line = LOESS trend",
    x = NULL, y = "AQI level"
  ) +
  theme_minimal()
```

## Autocorrelation

### ACF of AQI with all components (2017–2024)

```{r}
complete_data %>%
  as_tibble() %>%
  select(Start, AQI) %>%
  as_tsibble(index = Start) %>%
  ACF(AQI, lag_max = 48 * 7) %>% # up to one week of hourly lags
  autoplot() +
  labs(title = "ACF of AQI (2017–2024)", x = "Lag (hours)", y = "ACF") +
  theme_minimal()
```

### ACF of AQI noise - 1 year seasonality removed (STL remainder)

```{r}
noise_season_year %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise (STL Remainder) - 1 year season", x = "Lag (hours)", y = "ACF") +
  theme_minimal()
```

### ACF after removing annual + weekly + daily seasonality

```{r}
noise_full <- complete_data %>%
  model(STL(AQI ~ trend() + season("1 year") + season("week") + season("day"), robust = TRUE)) %>%
  components()

noise_full %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise (STL: annual + weekly + daily removed)", x = "Lag (hours)", y = "ACF") +
  theme_minimal()
```

### Stationarity tests
```{r}
# tu asi este pridat ADF test

# adf

complete_data %>%
  as_tibble() %>%
  select(Start, AQI) %>%
  as_tsibble(index = Start) %>%
  features(AQI, unitroot_kpss)
```