---
title: "AQI Prediction - Elements of AI Capstone Project"
author: "Filip Škrabák & Patrik Prizbul"
date: "2026-03-15"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(arrow)
library(lubridate)
library(tsibble)
library(here) # for handling paths correctly
library(corrplot)
library(zoo) # for na.approx
library(forecast)
library(generics)
library(fable)
library(feasts)
library(ggtime) ## gg_season
library(tseries) # for adf.test
library(MASS, exclude = "select") #ordinal regression polr()
here::i_am("project.Rmd")
```

# Introduction

The main objective of this project is to predict the Air Quality Index (AQI) for Bratislava, Slovakia, using historical air quality and weather data from Mamateyova street.
This requires us to merge two datasets - the pollutants dataset and the weather dataset. This is a multi-class classification problem, since AQI is categorized into 6 classes, from "Good" to "Extremely poor".

## Project Hypothesis: (concept)

Recent pollutant concentrations and weather conditions in Bratislava are jointly sufficient to predict the AQI 24 hours ahead.

# Pollutants Dataset

-   This dataset comes from [European Enviroment Agency (EEA) Air Quality Download Service](https://eeadmz1-downloads-webapp.azurewebsites.net/)
-   It contains pollutants from year 2017 until the end of 2024
-   Dataset type is Primary Validated Data (E1a), which is verified
-   Timezone is stated as UTC+1

First, we load the dataset and take a look at its structure.

```{r}
pollutants_data <- open_dataset(here("pollutants"))
glimpse(pollutants_data)
```

There are 12 columns and 902 260 rows.
First, lets keep only the concerned station (SK0001A), the years of interest (2017-2024) and relevant columns.

```{r}
filtered_air_data <- pollutants_data %>%
  select(Start, Pollutant, Value, Samplingpoint, Validity) %>% # lazy loading
  filter(grepl("SK0001A", Samplingpoint)) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024) %>%
  collect() 
```

Lets look at numeric ranges and unique values of the columns.

```{r}
summary(filtered_air_data)
```

The summary suggests that: - The `Value` column contains some negative values, which are probably sensor errors.
- The `Validity` is either 1 or 2.
According to data source, both are valid, with 2 being below detection limit.
We are keeping both.

Next, we should translate the `Pollutant` codes into human readable names and reshape the data to have columns for each pollutant.
Finally, we make sure the data is sorted by measurement time and we convert it into a tsibble for time series analysis.

```{r}
clean_air_data <- filtered_air_data %>%
  # Translate EU codes into factors
  mutate(Pollutant_Name = case_when(
    Pollutant == 1 ~ "SO2",
    Pollutant == 5 ~ "PM10",
    Pollutant == 7 ~ "O3",
    Pollutant == 8 ~ "NO2",
    #Pollutant == 10 ~ "CO",
    Pollutant == 6001 ~ "PM2.5"
  )) %>%
  select(Start, Pollutant_Name, Value) %>%
  pivot_wider(
    names_from = Pollutant_Name, 
    values_from = Value
  ) %>%
  arrange(Start) %>%
  as_tsibble(index = Start)

glimpse(clean_air_data)
#View(clean_air_data)
```

Now we have a cleaned dataset with 5 pollutants.
Lets check whether our time series has any gaps.
Our data is hourly.

Lets check gaps in our timeseries whether there are any missing hours.
For now, we will

```{r}
gaps <- clean_air_data %>%
  scan_gaps()

nrow(gaps)
gaps
```

There are 410 gaps.
For now, lets fill them with NA values.
Also check the tsibble gaps again to make sure they are filled.

```{r}
clean_air_data <- clean_air_data %>%
  fill_gaps()

has_gaps(clean_air_data)
interval(clean_air_data)
duplicates(clean_air_data)
```

The output confirms there are no more gaps, the interval is 1 hour and there are no duplicates.

```{r}
complete_rows <- sum(complete.cases(clean_air_data)) # 60845
rows <- nrow(clean_air_data) # 70128

missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage

# longest gap
clean_air_data %>%
  as_tibble() %>%
  pivot_longer(-Start, names_to = "pollutant", values_to = "value") %>%
  arrange(pollutant, Start) %>%
  group_by(pollutant) %>%
  mutate(is_na = is.na(value),
         group = cumsum(!is_na & !lag(!is_na, default = TRUE))) %>%
  filter(is_na) %>%
  group_by(pollutant, group) %>%
  summarise(gap_start = min(Start), gap_end = max(Start), hours = n(), .groups = "drop") %>%
  arrange(desc(hours)) %>%
  select(-group) %>%
  mutate(days = hours / 24)

```

13.24% of our data is missing.
The longest gap (sequence of NA values) for any pollutant is 10.2 days for NO2.

We have dicided that impute the gaps with linear interpolation for small gaps (3 hours). 
Longer gaps will be kept as NA, as replacing them could disrupt the seasonality or trends in data. 
The 3 hour gaps should be short enough to not cause significant bias.

```{r}
# @TODO: Old code, unused currently (might be used later if needed)
# We use a 2 step approach.
# Linear interpolation for small gaps (smaller than 8 hours) - Mean for same hour of day - for larger gaps, this respects the cycles in the data
impute_missing_values <- function(x, timestamps, max_linear = 8, window_days = 3) {
  x_out <- zoo::na.approx(x, na.rm = FALSE, maxgap = max_linear)

  hrs <- lubridate::hour(timestamps)
  n   <- length(x_out)

  global_hour_mean <- tapply(x, hrs, mean, na.rm = TRUE)

  for (i in seq_len(n)) {
    if (!is.na(x_out[i])) next

    neighbours <- max(1, i - window_days * 24) : min(n, i + window_days * 24)

    same_hour_and_known <- neighbours[
      hrs[neighbours] == hrs[i] & !is.na(x_out[neighbours])
    ]

    x_out[i] <- if (length(same_hour_and_known) > 0) {
      mean(x_out[same_hour_and_known]) 
    } else {
      global_hour_mean[[ as.character(hrs[i]) ]] 
    }
  }
  x_out
}
# END old code

complete_air_data <- clean_air_data %>%
  mutate(across(c(SO2, PM10, O3, NO2, `PM2.5`), ~ 
    zoo::na.approx(., na.rm = FALSE, maxgap = 3)
      #impute_missing_values(.x, Start)
    ))

complete_rows <- sum(complete.cases(complete_air_data)) # 60845
rows <- nrow(complete_air_data) # 70128
missing_percentage <- (rows - complete_rows) / rows * 100
missing_percentage
```

Now, after running the linear interpolation for maximum of 3 hour gaps, we reduced the missing percentage to 3.94%, which is acceptable.

# Weather Dataset

-   This dataset comes from [Open Meteo Historical Weather](https://open-meteo.com/en/docs/historical-weather-api)
-   It contains weather data from year 2017 until the end of 2024
-   Timezone is stated as GMT+0

We now load the dataset and rename the columns.

```{r}
weather_file_path <- here("weather", "open-meteo-48.12N17.10E163m.csv")

weather_raw <- read_csv(weather_file_path, skip = 3) %>%
  # We rename the columns manually, as the original names contain spaces and units
  rename(
    Start                 = time,
    Temperature           = `temperature_2m (°C)`,
    Humidity              = `relative_humidity_2m (%)`,
    Dew_Point             = `dew_point_2m (°C)`,
    Apparent_Temp         = `apparent_temperature (°C)`,
    Rain                  = `rain (mm)`,
    Precipitation         = `precipitation (mm)`,
    Snow_Depth            = `snow_depth (m)`,
    Snowfall              = `snowfall (cm)`,
    Weather_Code          = `weather_code (wmo code)`,
    Pressure_MSL          = `pressure_msl (hPa)`,
    Cloud_Cover           = `cloud_cover (%)`,
    Surface_Pressure      = `surface_pressure (hPa)`,
    Cloud_Cover_Low       = `cloud_cover_low (%)`,
    Cloud_Cover_Mid       = `cloud_cover_mid (%)`,
    Cloud_Cover_High      = `cloud_cover_high (%)`,
    ET0                   = `et0_fao_evapotranspiration (mm)`,
    Vapour_Pressure_Def   = `vapour_pressure_deficit (kPa)`,
    Wind_Speed            = `wind_speed_10m (km/h)`,
    Wind_Speed_100m       = `wind_speed_100m (km/h)`,
    Wind_Direction        = `wind_direction_10m (°)`,
    Wind_Gusts            = `wind_gusts_10m (km/h)`,
    Wind_Direction_100m   = `wind_direction_100m (°)`,
    Soil_Temp_0_7         = `soil_temperature_0_to_7cm (°C)`,
    Soil_Temp_7_28        = `soil_temperature_7_to_28cm (°C)`,
    Soil_Temp_28_100      = `soil_temperature_28_to_100cm (°C)`,
    Soil_Temp_100_255     = `soil_temperature_100_to_255cm (°C)`,
    Soil_Moisture_0_7     = `soil_moisture_0_to_7cm (m³/m³)`,
    Soil_Moisture_7_28    = `soil_moisture_7_to_28cm (m³/m³)`,
    Soil_Moisture_28_100  = `soil_moisture_28_to_100cm (m³/m³)`,
    Soil_Moisture_100_255 = `soil_moisture_100_to_255cm (m³/m³)`
  ) %>%
  filter(year(Start) >= 2017, year(Start) <= 2024)

glimpse(weather_raw)
```

```{r}
cat("Rows:", nrow(weather_raw), "| Missing:", sum(!complete.cases(weather_raw)), "\n")
cat("Columns:", ncol(weather_raw), "\n")
```

## Data Cleaning

The weather dataset has no missing values. It has 31 columns. 

We now try to identify collinearity (or linear relationships) between variables by looking at the correlation heatmap.
The heatmap is hierarchically clustered, so the most correlated variables are grouped together, which helps us identify clusters of collinear features.
```{r fig.height=8, fig.width=8}
cor_matrix <- weather_raw %>%
  select(-Start, -Weather_Code) %>%
  cor(use = "complete.obs")

cor_matrix %>%
  corrplot(
    method = "color", type = "upper", order = "hclust",
    addCoef.col = "black",
    number.cex = 0.4,
    number.digits = 1,
    tl.cex = 0.55, tl.col = "black",
    col = colorRampPalette(c("steelblue", "white", "firebrick"))(200),
    title = "Weather variable correlations",
    mar = c(0, 0, 1.5, 0)
  )
```


```{r}
cor_matrix %>%
  as_tibble(rownames = "Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation") %>%
  filter(Var1 < Var2) %>%
  arrange(desc(abs(Correlation))) %>%
  head(20)

```

The correlation heatmap confirms that we some variables are extremely highly correlated. Some of there are caused by measuring the same underlying phenomenon, but at different heights (e.g. wind speed at 10m and 100m), or they are calculated from the other columns (e.g. apparent temperature is calculated from temperature, humidity, and wind)

```{r}
weather_data <- weather_raw %>%
  select(-c(
    Wind_Speed_100m, Wind_Direction_100m, 
    Soil_Temp_7_28, Soil_Temp_28_100, Soil_Temp_100_255,
    Soil_Moisture_7_28, Soil_Moisture_28_100, Soil_Moisture_100_255,
    Surface_Pressure, Apparent_Temp, Snow_Depth, ET0
  ))
```

# Merging the datasets

```{r}
merged_data <- complete_air_data %>%
  inner_join(weather_data, by = "Start")

glimpse(merged_data)
summary(merged_data)
has_gaps(merged_data)
```

After the redundant variables are removed, we retain 24 features in total.
The merge has not changed the row count, and our time series has no gaps.

# AQI Calculation

Next step is to add AQI. AQI is calculated based on the worst pollutant in the area, so we calculate AQI level for each pollutant and then take the maximum of those as the overall AQI level. There are 6 AQI levels, from 1 (good) to 6 (extremely poor), based on the concentration thresholds defined by the European Environment Agency (EEA).

```{r}

# AQI levels and colors
aqi_levels <- c("Good", "Fair", "Moderate", "Poor", "Very poor", "Extremely poor")
aqi_colors <- c("Good"           = "#50f0e6",
                "Fair"           = "#50ccaa",
                "Moderate"       = "#f0e641",
                "Poor"           = "#ff5050",
                "Very poor"      = "#960032",
                "Extremely poor" = "#7d2181")

pollutant_colors <- c(
  "PM2.5"       = "#960032",
  "PM10"        = "#ff5050",
  "NO2"         = "#f0e641",
  "O3"          = "#50ccaa",
  "SO2"         = "#50f0e6",
  "Tie/Missing" = "#cccccc"
)

# European AQI calculation
complete_data_six <- merged_data %>%
  mutate(
    AQI_PM2.5 = case_when(
      `PM2.5` <=   5 ~ 1L, `PM2.5` <=  15 ~ 2L, `PM2.5` <=  50 ~ 3L,
      `PM2.5` <=  90 ~ 4L, `PM2.5` <= 140 ~ 5L, `PM2.5` >  140 ~ 6L
    ),
    AQI_PM10 = case_when(
      PM10 <=  15 ~ 1L, PM10 <=  45 ~ 2L, PM10 <= 120 ~ 3L,
      PM10 <= 195 ~ 4L, PM10 <= 270 ~ 5L, PM10 >  270 ~ 6L
    ),
    AQI_O3 = case_when(
      O3 <=  60 ~ 1L, O3 <= 100 ~ 2L, O3 <= 120 ~ 3L,
      O3 <= 160 ~ 4L, O3 <= 180 ~ 5L, O3 >  180 ~ 6L
    ),
    AQI_NO2 = case_when(
      NO2 <=  10 ~ 1L, NO2 <=  25 ~ 2L, NO2 <=  60 ~ 3L,
      NO2 <= 100 ~ 4L, NO2 <= 150 ~ 5L, NO2 >  150 ~ 6L
    ),
    AQI_SO2 = case_when(
      SO2 <=  20 ~ 1L, SO2 <=  40 ~ 2L, SO2 <= 125 ~ 3L,
      SO2 <= 190 ~ 4L, SO2 <= 275 ~ 5L, SO2 >  275 ~ 6L
    ),
    AQI = pmax(AQI_PM2.5, AQI_PM10, AQI_O3, AQI_NO2, AQI_SO2, na.rm = TRUE),
    AQI_Label = factor(
      case_when(
        AQI == 1L ~ aqi_levels[1],
        AQI == 2L ~ aqi_levels[2],
        AQI == 3L ~ aqi_levels[3],
        AQI == 4L ~ aqi_levels[4],
        AQI == 5L ~ aqi_levels[5],
        AQI == 6L ~ aqi_levels[6],
      ),
      levels = aqi_levels,
      ordered = TRUE # AQI is an ordered factor
    )
  )

glimpse(complete_data_six)
```

# Exploratory Data Analysis (EDA)

```{r}
theme_set(theme_minimal() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"))) 
```

## AQI Distribution

```{r}
complete_data_six %>%
  count(AQI_Label) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ggplot(aes(AQI_Label, n, fill = AQI_Label)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", percent)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = aqi_colors) +
  labs(title = "AQI Distribution (2017-2024)", x = "AQI", y = "Count")
```

Air quality in Bratislava is mostly Fair and Moderate (~89%). Good air quality happens in 5.2% of hours. Poor (original EEA classes 4-6, including Very poor and Extremely poor) accounts for 5.7% of hours.

The dataset is imbalanced due to Bratislava having Fair and Moderate air quality most of the time. The Poor class accounts for only 5.7% of the data, yet it is a critical class to be able to predict.

## Merging Rare AQI Classes

Since Very poor (0.3%) and Extremely poor (0.1%) are extremely rare in the dataset, and health warning for general population is issued by Poor level, we merge classes 4 (Poor), 5 (Very poor), and 6 (Extremely poor) into a single "Poor" class. This reduces the problem from 6 to 4 classes: Good, Fair, Moderate, Poor.

```{r}
# Redefine aqi_levels and aqi_colors for the 4 classes
aqi_levels <- c("Good", "Fair", "Moderate", "Poor")
aqi_colors <- c("Good"     = "#50f0e6",
                "Fair"     = "#50ccaa",
                "Moderate" = "#f0e641",
                "Poor"     = "#ff5050")

# Remap AQI
complete_data <- complete_data_six %>%
  mutate(
    AQI = pmin(AQI, 4L),
    AQI_Label = factor(aqi_levels[AQI], levels = aqi_levels, ordered = TRUE)
  )

complete_data %>% 
  as_tibble() %>% 
  count(AQI, AQI_Label) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

The merged "Poor" class accounts for about 5.7% of the data, which is still a minority, but large enough for the models to learn from. We also preserve the health related distinction.
Classes 1 to 3 represent safe conditions, while class 4 signals actionable poor air quality.

## AQI Drivers

As AQI is calculated based on the worst pollutant in the area, we should find out which is the most influencing one. When there are multiple pollutants with the same AQI level, we will label it as "Tie".

```{r}
aqi_drivers <- complete_data %>%
  as_tibble() %>%
  filter(!is.na(AQI)) %>%
  filter(AQI > 1) %>% # "Good" class is the baseline
  mutate(
    n_tied = (AQI_PM2.5 == AQI) + (AQI_PM10 == AQI) +
             (AQI_O3    == AQI) + (AQI_NO2  == AQI) +
             (AQI_SO2   == AQI),
    driver = case_when(
      is.na(n_tied) | n_tied != 1 ~ "Tie/Missing",
      AQI_PM2.5 == AQI ~ "PM2.5",
      AQI_PM10  == AQI ~ "PM10",
      AQI_O3    == AQI ~ "O3",
      AQI_NO2   == AQI ~ "NO2",
      AQI_SO2   == AQI ~ "SO2"
    )
  )

aqi_drivers %>%
  count(driver) %>%
  mutate(Percentage = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n)) %>%
  rename(Pollutant = driver, Hours = n)
```

From the table, we can see that PM2.5 influences the AQI the most, in **20.3%** of cases. The second most influencing pollutant, NO2 (18.2%), is caused by traffic emissions, and the third one, O3 (11.3%), peaks in summer months, as it is formed by photochemical reactions in the presence of sunlight. 

```{r}
aqi_drivers %>%
  summarise(n = n(), .by = c(AQI, driver)) %>%
  mutate(pct = 100 * n / sum(n), .by = AQI) %>%
  mutate(
    driver = factor(driver, levels = c("PM2.5", "PM10", "NO2", "O3", "SO2", "Tie/Missing")),
    AQI    = factor(AQI, levels = 1:4, labels = aqi_levels)
  ) %>%
  ggplot(aes(AQI, pct, fill = driver)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = if_else(pct >= 5, sprintf("%.0f%%", pct), "")),
            position = position_stack(vjust = 0.5), size = 3.2, color = "white", fontface = "bold") +
  scale_fill_manual(values = pollutant_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), expand = c(0, 0)) +
  labs(
    title = "Which pollutant drives the AQI? (by class)",
    x = "AQI class", y = "Share of hours (%)", fill = "Pollutant"
  )
```

The Poor class (class 4) is dominated by O3 (37%) and NO2 (31%), with PM2.5 contributing 21%. This is consistent with summer ozone peaks and traffic-related nitrogen dioxide (NO2). 

## Weekly average of AQI across years

```{r}
complete_data %>%
  as_tibble() %>%
  mutate(
    Week = floor_date(Start, "week")
  ) %>%
  group_by(Week) %>%
  summarise(AQI = mean(AQI, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(Week, AQI)) +
  geom_line(color = "steelblue", linewidth = 0.4, alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "firebrick", linewidth = 1) +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    breaks = 1:4,
    labels = aqi_levels,
    limits = c(1, 4)
  ) +
  labs(
    title = "Weekly average AQI (2017-2024)",
    subtitle = "Red line = LOESS trend",
    x = NULL, y = "AQI level"
  )
```

The weekly average plot shows a clear repeating seasonal pattern. AQI rises during winter months, and falls during spring and summer. LOESS trend reveals a slight long term improvement in air quality. 

At the beginning of 2017, there is a spike in AQI. This was caused by extremely cold January, which led to increased heating emissions [1].

## Outliers in data

Boxplots of pollutant concentrations (µg/m³) show which pollutants drive extreme AQI values.

```{r}
complete_data %>%
  as_tibble() %>%
  select(Start, PM2.5, PM10, NO2, O3, SO2) %>%
  pivot_longer(-Start, names_to = "Pollutant", values_to = "Concentration") %>%
  mutate(Pollutant = factor(Pollutant, levels = c("PM2.5", "PM10", "NO2", "O3", "SO2"))) %>%
  ggplot(aes(Pollutant, Concentration, fill = Pollutant)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3, linewidth = 0.4) +
  facet_wrap(~Pollutant, scales = "free_y", nrow = 1) +
  labs(title = "Pollutant distributions with outliers (2017-2024)", x = NULL, y = "Concentration (µg/m³)") +
  theme(legend.position = "none", axis.text.x = element_blank())
```

The plots show that NO2 and PM2.5 have the most frequent outlier hours. SO2, on the other hands, shows the most extreme outliers. As those readings are real world, validated sensor measurements, not errors, they are kept in the dataset, as they represent the most critical air quality events, which are important to predict.

## Decomposition and seasonality

STL decomposition of AQI:

```{r}
complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  model(STL(AQI ~ season(365*24) + season(7*24) + season(24), robust = TRUE)) %>%
  components() %>%
  autoplot(linewidth = 0.001) +
  labs(title = "STL Decomposition of AQI", x = "Time", y = "AQI") + 
  theme(panel.background = element_rect(),
        panel.grid.major = element_line())

data_season_year <- complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  model(STL(AQI ~ trend() + season("1 year"), robust = TRUE)) %>%
  components()

data_season_year_day <- complete_data %>%
  mutate(AQI = na.interp(AQI)) %>%
  model(STL(AQI ~ trend() + season("1 year") + season("day"), robust = TRUE)) %>%
  components()

```
@TODO: opis dekompozicie

```{r}
monthly_aqi <- complete_data %>%
  index_by(Month = yearquarter(Start)) %>%
  summarise(AQI = mean(AQI))

monthly_aqi %>%
  gg_season(AQI, labels = "right") +
  scale_y_continuous(breaks = 1:4,
                     labels = aqi_levels,
                     limits = c(1, 4)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017-2024)")

monthly_aqi %>%
  gg_subseries(AQI) +
  scale_y_continuous(breaks = 1:4,
                     labels = aqi_levels,
                     limits = c(1, 4)) +
  labs(y = "Mean Monthly AQI", x = "Month",
       title = "Seasonal plot - monthly AQI by year (2017-2024)")

complete_data %>%
  gg_season(PM2.5, labels = "right") +
  labs(y = "PM2.5", x = element_blank(),
       title = "Yearly PM2.5 (2017-2024)")


```

The plot shows that AQI is worse around winter months (November to March), spring and autumn months have the best air quality, while summer is somewhere in the middle.
This is likely caused by heating in the winter months (by PM2.5).

```{r}
hourly_aqi <- complete_data %>%
  as_tibble() %>%
  mutate(Hour = hour(Start)) %>%
  group_by(Hour) %>%
  summarise(
    Mean  = mean(AQI, na.rm = TRUE),
    Q25   = quantile(AQI, 0.25, na.rm = TRUE),
    Q75   = quantile(AQI, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

hourly_aqi %>%
ggplot(aes(Hour, Mean)) +
  geom_line(color = "steelblue", linewidth = 1) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(
    breaks = 1:4,
    labels = aqi_levels,
    limits = c(1, 4)
  ) +
  labs(
    title = "Hourly AQI pattern (hour of day, 2017-2024)",
    x = element_blank(), y = element_blank()
  )
```

The hourly plot shows that there is a slight peak at around 6am, and an evening peak at around 16pm, or 22pm, likely due to traffic and heating patterns.


## Autocorrelation

This step should help us understand which lags are the most important for predicting AQI.

### ACF of AQI with all components (2017-2024)

```{r}
complete_data %>%
  ACF(AQI, lag_max = 48 * 7) %>% # up to one week of hourly lags
  autoplot() +
  labs(title = "ACF of AQI (2017-2024)", x = "Lag (hours)", y = "ACF")
```

The plot shows that AQI lags slowly decay. Spikes at 24 hour cycles confirm this daily cycle. There is also a spike at lag 168 (7 days).

### ACF of AQI noise - 1 year seasonality removed (STL remainder)

```{r}
data_season_year %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise - 1 year season", x = "Lag (hours)", y = "ACF")
```

By removing the 1 year seasonality, the long-range slow decay practically disappears, and the 24 hour spikes remain prominent. 

### ACF after removing annual + daily seasonality

```{r}
data_season_year_day %>%
  ACF(remainder, lag_max = 48 * 7) %>%
  autoplot() +
  labs(title = "ACF of AQI Noise (STL: annual + daily removed)", x = "Lag (hours)", y = "ACF")
```

Now that both annual and daily components are removed, the lags decay in a much smoother manner. The lag 1 stays high, but the 24 hour spike is gone compared to previous plots. 

### PACF of AQI

```{r}
complete_data %>%
  PACF(AQI, lag_max = 48) %>%
  autoplot() +
  labs(title = "PACF of AQI (2017-2024)", x = "Lag (hours)", y = "PACF")
```

The PACF plot confirms that the lags 1 and 2 are the most important. 
@TODO: mozno lepsi popis? toto je dobre pre ARIMU, mozno dat uplne prec ak sa nepouzije...

### Stationarity tests

We run the KPSS and ADF tests to check whether our series is stationary.

```{r warning=FALSE}
tseries::kpss.test(complete_data$AQI)
#tseries::kpss.test(noise_season_year$remainder)
#tseries::kpss.test(noise_season_year_day$remainder)
```

The KPSS rejects stationarity (H0: stationary), p value \< 0.01

```{r warning=FALSE}
tseries::adf.test(na.omit(complete_data$AQI))
```

ADF rejects the unit root (H0: series has a unit root), p value \< 0.01

This should indicate seasonal non-stationarity (trend-stationarity).
The mean is not constant, due to winter peaks and summer troughs, but there is no drift in the long run.
Since this time series does not have a unit root, we will not be using differencing as that would likely lead to overdifferencing.

## Predictor Correlations

To see possible multicollinearity and relationships between main predictors, we use the pairs plot.
A random sample of 3 000 rows is used (due to performance).

```{r fig.height=10, fig.width=10, warning=FALSE}
# correlation on upper panels
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r     <- cor(x, y, use = "complete.obs")
  r_abs <- abs(r)
  txt   <- format(round(r, digits), nsmall = digits)
  txt   <- paste0(prefix, txt)
  if (missing(cex.cor)) cex.cor <- 0.8 / strwidth(txt)
  col <- if (r >= 0) "firebrick" else "steelblue"
  text(0.5, 0.5, txt, cex = cex.cor * (0.5 + r_abs) / 1.5, col = col)
}

# histogram on the diagonal
panel.hist <- function(x, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5))
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y / max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "steelblue", border = "white")
}

set.seed(42)
pairs_data <- complete_data %>%
  as_tibble() %>%
  rename(PM25 = `PM2.5`) %>% # for clarity
  select(PM25, PM10, NO2, O3, SO2, AQI,
         Temperature, Humidity, Wind_Speed, Pressure_MSL) %>%
  drop_na() %>%
  slice_sample(n = 3000)

pairs(pairs_data,
      lower.panel = panel.smooth,
      upper.panel = panel.cor,
      diag.panel  = panel.hist,
      main = "Pred. relationships and correlations (n = 3 000)")
```

The pairs plot reveals several notable patterns.
PM2.5 and PM10 are strongly correlated (r = 0.85), which indicates that they largely capture the same signal.
NO2 is moderately correlated with PM2.5 and PM10 (r = 0.41), as they all peak during cold, stagnant winter episodes when heating emissions are high and dispersion is low.
O3 is negatively correlated with NO2 (r = -0.57), and does not show a linear relationship with PM2.5/PM10. O3 is also nearly uncorrelated with AQI (r = -0.04) as it is only the main driver in around 11% of hours, and its relationship with AQI isn't linear.
Temperature is quite strongly positively correlated with O3 (r = 0.65) as hot, sunny conditions drive ozone formation. The AQI correlation is near zero (r = -0.03).
Humidity is strongly negatively correlated with O3 (r = -0.73). Dry and sunny conditions drive high ozone. AQI correlation is near zero (r = -0.04).
Wind speed is negatively correlated with AQI (r = -0.33) and with most pollutants, as higher wind disperses pollution.
Pressure is weakly positively correlated with AQI (r = 0.13).


# Feature Engineering

Now, we transform the time series data into a format suitable for machine learning models - we basically make it a supervised learning problem, where the target variable is AQI 24 hours ahead (AQI_p24), and the features are the current and past values of AQI, pollutants, weather variables, and time features (hour, month). This allows us to use many different types of ML models. We decided the lag variables based on autocorrelation analysis.

```{r}
model_features <- complete_data %>%
  as_tibble() %>%
  arrange(Start) %>%
  transmute(
    Start,
    # Target variable
    AQI_p24 = lead(AQI, 24),
    # Autocorrelation lags
    AQI_t   = AQI,
    AQI_t1  = lag(AQI, 1),
    AQI_t2  = lag(AQI, 2),
    AQI_t24 = lag(AQI, 24),
    # Current pollutants
    PM25    = `PM2.5`,
    PM10    = PM10,
    NO2     = NO2,
    O3      = O3,
    SO2     = SO2,
    # Time
    hour    = hour(Start),
    month   = month(Start),
    # Weather
    across(c(Temperature, Humidity, Dew_Point, Rain, Precipitation,
             Snowfall, Weather_Code, Pressure_MSL, Cloud_Cover,
             Cloud_Cover_Low, Cloud_Cover_Mid, Cloud_Cover_High,
             Vapour_Pressure_Def, Wind_Speed, Wind_Direction,
             Wind_Gusts, Soil_Temp_0_7, Soil_Moisture_0_7))
  ) %>%
  drop_na() # remove lag NAs


# No weather variables (approach 1)
PARTITION_A <- c("AQI_t","AQI_t1","AQI_t2","AQI_t24",
                 "PM25","PM10","NO2","O3","SO2","hour","month")

# Maybe use partitioning method for this (Task 3)? (approach 2)
PARTITION_B <- model_features %>%
  select(-Start, -AQI_p24) %>%
  names()

```

# Modelling

All models are trained on 2017-2022 data and evaluated on 2023-2024.
This split ensures no future observations leak into the training data.

## Data Budgeting

```{r}
# Split model_features into train / test by year.
train_data <- model_features %>%
  filter(year(Start) <= 2022) %>%
  select(-Start) %>%
  # AQI_p24 is converted to an ordered factor (1-4) for classification.
  mutate(AQI_p24 = factor(AQI_p24, levels = 1:4, ordered = TRUE))

test_data <- model_features %>%
  filter(year(Start) >= 2023) %>%
  select(-Start) %>%
  mutate(AQI_p24 = factor(AQI_p24, levels = 1:4, ordered = TRUE))

cat("Train rows:", nrow(train_data), " | Test rows:", nrow(test_data), "\n")
```

```{r}
# Class distribution per split (imbalance check)
bind_rows(
  train_data %>% count(AQI_p24) %>% mutate(Split = "Train", pct = round(100 * n / sum(n), 1)),
  test_data  %>% count(AQI_p24) %>% mutate(Split = "Test",  pct = round(100 * n / sum(n), 1))
) %>%
  mutate(Label = aqi_levels[as.integer(AQI_p24)]) %>%
  select(Split, Class = AQI_p24, Label, Count = n, `Pct (%)` = pct) %>%
  arrange(Split, Class)
```

Class 4 (Poor) makes up 6.5% of the training data and 3.3% of the test data.
All four classes are represented, but Poor is a minority class.
We report per-class recall in every evaluation to make this limitation transparent.

## Scenario 1 (Three methods across two feature approaches)

Each method is evaluated on two feature sets:

- **Partition A** - AQI lags + current pollutants + time 
- **Partition B** - All features including weather variables

Comparing the two partitions tests whether current weather conditions add predictive value beyond pollutant lags and time features.

### Random Forest

```{r}
library(ranger)

RF_NUM_TREES <- 500 # tested: 100, 300, 500 — diminishing returns above 500
RF_MTRY <- 5 # features tried at each split, tried {4,5,6,8,10,15}, sqrt(p)=5 was optimal
RF_MIN_NODE_SIZE <- 5 # min samples in a leaf, tried {1,3,5,10,20}, 5 balances depth vs minority recall
RF_MAX_DEPTH <- NULL  # NULL = fully grown trees
RF_N_THREADS <- max(1, parallel::detectCores() - 1)
RF_SEED <- 2026
RF_USE_CLASS_WEIGHTS <- TRUE # true - use inverse weights
RF_WEIGHT_POWER <- 1.1 # exponent on inverse-frequency weights, tried {0.5,0.6,...,1.3}
# 1.1: best Poor recall/ accuracy tradeoff
```

```{r}
# Computes inverse-frequency case weights for a training set (equivalent to sklearn class_weight="balanced").
# Each sample is weighted by (1 / count_of_its_class)^power.
# power>1 amplifies the boost for rare classes.
compute_class_weights <- function(train_data, power = RF_WEIGHT_POWER) {
  class_counts <- table(train_data$AQI_p24)
  weights <- (1 / as.numeric(class_counts))^power
  weights[as.integer(train_data$AQI_p24)]
}

# Fit RF
fit_rf <- function(train_data, partition,
                   num_trees = RF_NUM_TREES,
                   mtry = RF_MTRY,
                   min_node_size = RF_MIN_NODE_SIZE,
                   max_depth = RF_MAX_DEPTH,
                   n_threads = RF_N_THREADS,
                   seed = RF_SEED,
                   use_weights = RF_USE_CLASS_WEIGHTS,
                   weight_power = RF_WEIGHT_POWER) {

  train_subset <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  weights <- if (use_weights) compute_class_weights(train_subset, power = weight_power) else NULL
  ranger(
    AQI_p24 ~ .,
    data = train_subset,
    num.trees = num_trees,
    mtry = mtry,
    importance = "permutation",
    min.node.size = min_node_size,
    max.depth = max_depth,
    num.threads = n_threads,
    seed = seed,
    case.weights = weights
  )
}

# Returns integer vectors of actual and predicted classes for a fitted RF
predict_rf <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(test_data$AQI_p24)
  preds  <- as.integer(predict(model, data = test_features)$predictions)
  list(actual = actual, preds = preds)
}

# Computes overall test-set metrics for a fitted RF
eval_rf <- function(model, test_data, partition) {
  p      <- predict_rf(model, test_data, partition)
  actual <- p$actual
  preds  <- p$preds

  tibble(
    # We calculate RMSE/MAE because AQI is an ordered factor.
    RMSE = round(sqrt(mean((actual - preds)^2)), 4),
    MAE = round(mean(abs(actual - preds)), 4),
    `Acc (%)` = round(mean(preds == actual) * 100, 2),
    `+-1 Class Acc (%)` = round(mean(abs(preds - actual) <= 1) * 100, 2) # within 1 class
  )
}

# Per-class recall - fraction of each true class correctly predicted
# This helps us understand how well the model predicts individual AQI classes
class_recall_rf <- function(model, test_data, partition) {
  p      <- predict_rf(model, test_data, partition)
  actual <- p$actual
  preds  <- p$preds

  return(tibble(Class = 1:4, Label = aqi_levels) %>%
    mutate(
      Actual_n = map_int(Class, \(k) sum(actual == k)),
      Pred_n = map_int(Class, \(k) sum(preds  == k)),
      `Recall (%)` = map_dbl(Class, \(k) {
        n_actual <- sum(actual == k)
        if (n_actual == 0) NA_real_ else round(mean(preds[actual == k] == k) * 100, 1)
      })
    ))
}

# bar chart of top-20 permutation importance scores
plot_importance <- function(model, title = "RF Permutation Importance") {
  feat_and_imp <- tibble(
    Feature = names(model$variable.importance),
    Importance = model$variable.importance
  ) 
    
  feat_and_imp%>%
    arrange(desc(Importance)) %>%
    slice_head(n = 20) %>%
    mutate(Feature = fct_reorder(Feature, Importance)) %>%
    ggplot(aes(Importance, Feature)) +
    geom_col(fill = "steelblue") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    labs(title = title, x = "Permutation importance", y = NULL)
}
```

```{r}
rf_A <- fit_rf(train_data, PARTITION_A, use_weights = FALSE)
rf_B <- fit_rf(train_data, PARTITION_B, use_weights = FALSE)

rf_B_weighted <- fit_rf(train_data, PARTITION_B, use_weights = TRUE)
```

Overall performance on the 2023-2024 test set:

```{r}
evals_RF <- list(
  "A - pollutants only" = eval_rf(rf_A, test_data, PARTITION_A),
  "B - all features" = eval_rf(rf_B, test_data, PARTITION_B),
  "B - all features (weighted)" = eval_rf(rf_B_weighted, test_data, PARTITION_B)
) 

evals_RF %>%
  list_rbind(names_to = "Partition")
```

Per-class recall (how often each AQI class was correctly predicted):
```{r}
cat("--- Partition A ---\n")
class_recall_rf(rf_A, test_data, PARTITION_A) %>% 
  print()
cat("\n--- Partition B (unweighted) ---\n")
class_recall_rf(rf_B, test_data, PARTITION_B) %>% 
  print()
cat("\n--- Partition B (weighted) ---\n")
class_recall_rf(rf_B_weighted, test_data, PARTITION_B) %>% 
  print()
```

```{r}
plot_importance(rf_A, "RF Permutation Importance (Partition A)")
```

```{r}
plot_importance(rf_B_weighted, "RF Permutation Importance (Partition B weighted)")
```

#### Confusion matrix

```{r}
# We take predictions from the weighted model
preds_B_weighted <- predict_rf(rf_B_weighted, test_data, PARTITION_B)
actual_cls <- preds_B_weighted$actual
pred_cls <- preds_B_weighted$preds
act_pred_tb <- tibble(Actual = actual_cls, Predicted = pred_cls)

# row-normalised confusion df
act_pred_tb %>%
  mutate(
    Actual = factor(Actual, levels = 1:4, labels = aqi_levels),
    Predicted = factor(Predicted, levels = 1:4, labels = aqi_levels)
  ) %>%
  count(Actual, Predicted) %>%
  complete(Actual, Predicted, fill = list(n = 0)) %>%
  mutate(Recall = n / sum(n), .by = Actual) %>%
  ggplot(aes(Predicted, Actual, fill = Recall)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = if_else(n > 0, sprintf("%.0f%%\n(n=%d)", Recall * 100, n), "")),  
            size = 2.8, lineheight = 1.2) +
  scale_fill_gradient(low = "white", high = "steelblue",
                      labels = scales::percent, limits = c(0, 1)) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev(aqi_levels)) +
  labs(
    title = "Confusion Matrix for RF Partition B (weighted, pwr=1.1)",
    subtitle = "Each cell = % of true class predicted as that column",
    x = "Predicted class", y = "True (Actual) class", fill = "Recall"
  ) +
  theme_minimal() 
```

The confusion matrix shows recall per class on the diagonal.
The model is strong for Fair (76.9%) and Moderate (54.2%). With class weighting (power 1.1), Poor recall rises to **37.4%**, which is a meaningful improvement over the unweighted model (13%).
Most misclassifications are adjacent-class (+-1) errors (for example, Poor predicted as Moderate, or Good predicted as Fair).

#### Predicted vs. actual class distribution

```{r}
act_pred_tb %>%
  pivot_longer(c(Actual, Predicted), names_to = "Type", values_to = "Class") %>%
  count(Type, Class) %>%
  mutate(Label = factor(aqi_levels[Class], levels = aqi_levels)) %>%
  complete(Type, Label, fill = list(n = 0)) %>%
  ggplot(aes(Label, n, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Actual" = "steelblue", "Predicted" = "tomato")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Predicted vs. actual class distribution (test set 2023-2024)",
    x = NULL, y = "Count", fill = NULL
  ) +
  theme_minimal()
```

The model over-predicts Fair and under-predicts Good and Moderate.

#### One vs Rest ROC and AUC

One VS Rest ROC (OvR) allows us to see how tell the model separates each class from the others.
The `ranger` package doesn't provide probability estimates without `probability = TRUE`, so we opted to train a separate probability forest for this evaluation. All parameters and hyperparameters are kept the same as the original RF to ensure a fair comparison.

```{r}
# Probability forest
rf_B_prob <- ranger(
  AQI_p24 ~ .,
  data = train_data %>% select(all_of(c(PARTITION_B, "AQI_p24"))),
  num.trees = RF_NUM_TREES,
  mtry = RF_MTRY,
  min.node.size = RF_MIN_NODE_SIZE,
  num.threads = RF_N_THREADS,
  seed = RF_SEED,
  case.weights = compute_class_weights(train_data),
  probability = TRUE # outputs probabilities instead of class labels
)
```

```{r}
library(pROC)

prob_matrix <- predict(rf_B_prob, data = test_features_B)$predictions

# OVR ROC
roc_list <- map(1:4, function(k) { # loop over aqi
  binary <- as.integer(actual_cls == k) # 1 for class k, 0 for others
  if (length(unique(binary)) < 2) return(NULL)   # skip if no positives in test
  roc(binary, prob_matrix[, k], quiet = TRUE)
})

# tidy data frame of all ROC curves for plotting (for ggplot)
# bit more complex, but allows us to use more pretty ggplot with ordering
roc_df <- map(1:4, function(k) {
  if (is.null(roc_list[[k]])) return(NULL)
  coords_df <- pROC::coords(roc_list[[k]], "all", ret = c("fpr", "tpr")) # get FPR and TPR for all thresholds
  coords_df$class <- aqi_levels[k] # label for the class
  coords_df$auc <- round(pROC::auc(roc_list[[k]]), 3) # AUC for the class for legend
  coords_df
}) %>% list_rbind()

roc_df %>%
  mutate(label = paste0(class, "  (AUC=", auc, ")"), # label for the legend
         label = factor(label, levels = unique(label[order(auc, decreasing = TRUE)]))) %>%
  ggplot(aes(fpr, tpr, color = label)) +
  geom_line(linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_brewer(palette = "Dark2") + # color palette
  labs(
    title = "One VS Rest ROC (RF Partition B weighted)",
    x = "False Positive Rate", y = "True Positive Rate", color = "Class (AUC)"
  ) +
  theme_minimal()
```

The strongest class is **Poor (AUC = 0.889)**, meaning the model reliably separates poor air quality from safe conditions.
Good is also well separated (AUC = 0.776). Moderate and Fair, being the dominant classes, are harder to distinguish from each other: AUC = 0.728 and 0.700, respectively.

### Ordinal regression

```{r}
ORDINAL_WEIGHT_POWER <- 0.2 # 0 = no reweighting, 1 = fully balanced

# polr() treats weights as frequency weights (each row counts as w observations)
# Weights must sum to nrow(train_subset) so the effective sample size equals the actual sample size.
compute_class_weights_ordinal <- function(train_data, power = ORDINAL_WEIGHT_POWER) {
  class_counts <- table(train_data$AQI_p24)
  w_raw     <- (1 / as.numeric(class_counts))^power
  w_per_row <- w_raw[as.integer(as.character(train_data$AQI_p24))]
  # Normalize so weights sum to n (preserves effective sample size)
  w_per_row / sum(w_per_row) * nrow(train_data)
}

fit_ordinal <- function(train_data, partition) {
  train_subset <- train_data %>% select(all_of(c(partition, "AQI_p24")))
  polr(AQI_p24 ~ ., data = train_subset, Hess = TRUE,
       weights = compute_class_weights_ordinal(train_subset))
}

ordinal_A <- fit_ordinal(train_data, PARTITION_A)
ordinal_B <- fit_ordinal(train_data, PARTITION_B)
```

#### Coefficient table with p-values

`polr` only outputs t-values; p-values must be derived from the normal approximation (valid for large n).

```{r}
polr_summary_table <- function(model) {
  tbl <- as.data.frame(coef(summary(model)))
  tbl$p_value <- pnorm(abs(tbl[, "t value"]), lower.tail = FALSE) * 2
  round(tbl[, c("Value", "Std. Error", "t value", "p_value")], 4)
}

cat("=== Partition A ===\n"); print(polr_summary_table(ordinal_A))
cat("\n=== Partition B ===\n"); print(polr_summary_table(ordinal_B))
```

#### Goodness of fit - McFadden R2

```{r}
library(performance)

cat("Partition A:\n"); print(r2_mcfadden(ordinal_A))
cat("Partition B:\n"); print(r2_mcfadden(ordinal_B))
```

#### Performance evaluation

```{r}
eval_ordinal <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(as.character(test_data$AQI_p24))
  preds  <- as.integer(as.character(predict(model, newdata = test_features)))
  tibble(
    RMSE = round(sqrt(mean((actual - preds)^2)), 4),
    MAE  = round(mean(abs(actual - preds)), 4),
    `Acc (%)` = round(mean(preds == actual) * 100, 2),
    `+-1 Class Acc (%)` = round(mean(abs(preds - actual) <= 1) * 100, 2)
  )
}

class_recall_ordinal <- function(model, test_data, partition) {
  test_features <- test_data %>% select(all_of(partition))
  actual <- as.integer(as.character(test_data$AQI_p24))
  preds  <- as.integer(as.character(predict(model, newdata = test_features)))
  tibble(Class = 1:4, Label = aqi_levels) %>%
    mutate(
      Actual_n     = map_int(Class, \(k) sum(actual == k)),
      Pred_n       = map_int(Class, \(k) sum(preds  == k)),
      `Recall (%)` = map_dbl(Class, \(k) {
        n_actual <- sum(actual == k)
        if (n_actual == 0) NA_real_
        else round(mean(preds[actual == k] == k) * 100, 1)
      })
    )
}

list(
  "A - pollutants only" = eval_ordinal(ordinal_A, test_data, PARTITION_A),
  "B - all features"    = eval_ordinal(ordinal_B, test_data, PARTITION_B)
) %>%
  list_rbind(names_to = "Partition")

cat("--- Partition A ---\n")
class_recall_ordinal(ordinal_A, test_data, PARTITION_A) %>% print()
cat("\n--- Partition B ---\n")
class_recall_ordinal(ordinal_B, test_data, PARTITION_B) %>% print()
```

#### Assumption check: Proportional Odds (Brant test)

The core assumption of `polr` is proportional odds: every predictor must have the *same* log-odds effect at every cumulative cutpoint (parallel regression lines).

The `brant` function requires the training data to be in the calling environment, so we fit the model directly here (same parameters as `ordinal_A`).

```{r message=FALSE}
library(brant)

# brant() looks up the data object by name from the polr model call, so the data must exist as a named variable in the current environment.
train_subset_A_brant <- train_data %>% select(all_of(c(PARTITION_A, "AQI_p24")))
ordinal_A_brant <- polr(AQI_p24 ~ ., data = train_subset_A_brant, Hess = TRUE,
                        weights = compute_class_weights_ordinal(train_subset_A_brant))

cat("=== Brant test - Partition A ===\n")
print(brant(ordinal_A_brant))
```

H0: parallel regression assumption holds. Rejection (p < 0.05) means that predictor's effect differs across cutpoints. The PO assumption is violated.

#### Coefficient stability across cutpoints

Fitting three separate binary logistic models (one per cumulative cutpoint: <=1, <=2, <=3) and overlaying their coefficients shows *how* the effect changes across the scale. Flat lines support the PO assumption; sloping or crossing lines indicate violation.

```{r fig.height=6, fig.width=9}
train_for_binom <- train_data %>%
  select(all_of(c(PARTITION_A, "AQI_p24"))) %>%
  mutate(AQI_num = as.numeric(as.character(AQI_p24)))

cutpoint_coefs <- map_dfr(1:3, function(k) {
  fit <- glm(I(AQI_num <= k) ~ AQI_t + AQI_t1 + AQI_t2 + AQI_t24 +
               PM25 + PM10 + NO2 + O3 + SO2 + hour + month,
             data = train_for_binom, family = binomial)
  tibble(
    Cutpoint = paste0("<= ", k),
    Predictor = names(coef(fit))[-1],
    Coef = coef(fit)[-1]
  )
})

cutpoint_coefs %>%
  ggplot(aes(x = Cutpoint, y = Coef, group = Predictor)) +
  geom_line(linewidth = 0.7, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~Predictor, scales = "free_y", nrow = 3) +
  labs(
    title = "Proportional Odds check: coefficient stability across cutpoints",
    subtitle = "Flat lines = PO holds. Varying lines = PO violated.",
    x = "Cumulative cutpoint", y = "Logit coefficient"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

O3, NO2, PM25, hour, and month show clearly non-constant effects across cutpoints, confirming the Brant test. AQI_t1, AQI_t2, and PM10 are the only predictors for which PO holds. This means the standard `polr` model is misspecified for this data and a multinomial or partial-PO model would be more appropriate.

# References
- [1] [Air Quality in Slovakia - Ministry of Health](https://www.minzp.sk/files/oblasti/ovzdusie/ochrana-ovzdusia/dokumenty/strategia-ochrany-ovzdusia/hodnotenie_pre_strategiu_podklady.pdf)
- [2] [European Air Quality Index](https://www.eea.europa.eu/en/analysis/maps-and-charts/index)